### 2014-10-07 paper/paper.tex
Add template for paper

\documentclass{llncs}

\author{L. Thomas van Binsbergen\inst{1} \and Jeroen Bransen\inst{2} \and Koen Claessen\inst{3}}
\institute{todo, \email{todo@todo.co.uk}
\and Utrecht University, Utrecht, The Netherlands, \email{J.Bransen@uu.nl}
\and todo, \email{koen@chalmers.se}}
\title{SAT-solving Linear Ordered Attribute Grammars}

\begin{document}

\maketitle

\begin{abstract}
A nice abstract here.
\keywords{Attribute Grammars, SAT-Solving}
\end{abstract}

\section{Introduction}

\end{document}

### 2014-10-07 paper/formats.tex
Required latex files (including a messy format.tex)

% syntax highlighting
\newcommand{\mvar}[1]{$\;\texttt{\textcolor{black}{#1}}\;$}
\newcommand{\mkey}[1]{$\;\texttt{\textcolor{orange}{#1}}\;$}
\newcommand{\mmod}[1]{$\;\texttt{\textcolor{purple}{#1}}\;$}
\newcommand{\mdef}[1]{$\;\texttt{\textcolor{blue}{#1}}\;$}
\newcommand{\mcon}[1]{$\;\texttt{\textcolor{brickred}{#1}}\;$}
\newcommand{\mtype}[1]{$\;\texttt{\textcolor{brown}{#1}}\;$}
\newcommand{\mpar}[1]{$(#1)$}
\newcommand{\mnil}[1]{\mcon{[]}}
\newcommand{\cons}[3]{\mpar{\mval{#1}\mcon{#2}\mval{#3}}}
\newcommand{\ra}[0]{$\;\rightarrow\;$}
\newcommand{\Ra}[0]{$\;\Rightarrow\;$}
\newcommand{\ind}[0]{$\;\;\;\;$}

\newcommand{\plusplus}{\mathbin{+\!\!\!+}}

%\setlength\abovedisplayskip{4pt}
%\setlength\abovecaptionskip{0pt}
%\setlength\belowdisplayskip{0pt}
%\setlength\belowcaptionskip{0pt}
%\setlength\intextsep{4pt}

% grammars
\newcommand{\nt}[1]{\texttt{#1}}
\newcommand{\pr}[1]{\textit{\texttt{#1}}}
\newcommand{\tm}[1]{\underline{#1}}
\newcommand{\lf}[1]{\texttt{#1}.}
\newcommand{\rf}[1]{@\texttt{#1}.}
\newcommand{\f}[1]{@\texttt{#1}}
\newcommand{\ff}[1]{\texttt{#1}}

\newcommand{\app}[1]{\textsc{#1}}

% sets
\newcommand{\set}[2]{\{\;#1\;\;|\;\;#2\;\}}
\newcommand{\sand}[0]{,\;}

% quantification
\newcommand{\univ}[2]{\forall(#1)\;(#2)}
\newcommand{\exis}[2]{\exists(#1)\;(#2)}

% administration
\newcommand{\toindex}[1]{\emph{#1}\index{#1}}
\newcommand{\toindexwith}[2]{\emph{#1}\index{#2}}

% visit-sequences
\newcommand{\vsrelsize}{1.5cm}
\tikzstyle{vsstyle}=[shape=rectangle,draw]

\newcommand{\myvs}[3]{
\draw[style=vsstyle] (#1 *\vsrelsize,0) rectangle (#2 *\vsrelsize, 0.9cm) node[anchor=north east] (v1#1) {\begin{varwidth}{\vsrelsize}\small{#3}\end{varwidth}};
}

% agpictures
\definecolor{light-gray}{gray}{0.83}
\newdimen\agrelsize
\setlength{\agrelsize}{0.30cm}
\tikzstyle{agnontermstyle}=[shape=circle,draw,minimum size=\agrelsize]
\tikzstyle{agchldstyle}=[shape=circle,fill=red!70,minimum size=\agrelsize]
\tikzstyle{agattrstyle}=[shape=rectangle,draw,minimum size=\agrelsize,]
\tikzstyle{agdepstyle}=[black,>=triangle 60]
\tikzstyle{agidpstyle}=[gray,>=triangle 60]
\tikzstyle{inputstyle}=[shape=rectangle,fill=light-gray,draw,minimum size=\agrelsize]
\tikzstyle{outputstyle}=[shape=rectangle,draw,minimum size=\agrelsize]
\newcommand{\agcoordpary}{6}   %% y-coordinate of parent attributes
\newcommand{\agcoordchldy}{0}  %% y-coordinate of child attributes
\newcommand{\agcoordlocy}{3}   %% y-coordinate of local attributes
\newcommand{\agbeginprod}[5]{} %% you could begin a picture here
\newcommand{\agendprod}[5]{}   %% you could end a picture here
\newcommand{\agparent}[5]{ \draw (#4 \agrelsize,\agcoordpary \agrelsize) node [style=agnontermstyle] (#3) {}; \draw (#3.north) node [anchor=south] {\footnotesize{ #2 }}; \draw (#3.south) node [anchor=north] {\footnotesize{ #5 }}; }
\newcommand{\agparattr}[6]{ \draw (#4 \agrelsize,\agcoordpary \agrelsize) node [style=#6] (#3) {}; \draw (#3.south) node [anchor=south] {\tiny{#2}}; }
\newcommand{\agchldattr}[6]{ \draw (#4 \agrelsize,\agcoordchldy \agrelsize) node [style=#6] (#3) {}; \draw (#3.north) node [anchor=north] {\tiny{#2}}; }
\newcommand{\agchldnonterm}[5]{ \draw (#4 \agrelsize,\agcoordchldy \agrelsize) node [style=agnontermstyle] (#3) {}; \draw (#3.south) node [anchor=north] {\footnotesize{ #2:#5 }}; }
\newcommand{\agchldterm}[5]{ \draw (#4 \agrelsize,\agcoordchldy \agrelsize) node [style=agchldstyle] (#3) {}; \draw (#3.south) node [anchor=north] {\footnotesize{ #2 }}; }
\newcommand{\agsibline}[3]{ \draw (#2.east) -- (#3.west); }
\newcommand{\agdeplineupdown}[4]{ \draw[->,style=agdepstyle] (#2.north) .. controls +(0,1) and +(0,-1) .. (#3.south); }
\newcommand{\agdeplineupup}[5]{ \draw[->,style=agdepstyle] (#2.north) .. controls +(0,#5) and +(0,#5) .. (#3.north); }
\newcommand{\agidplineup}[5]{ \draw[->,style=agidpstyle] (#2.north) .. controls +(0,#5) and +(0,#5) .. (#3.north); }
\newcommand{\agidplineupdown}[5]{ \draw[->,style=agidpstyle] (#2.north) .. controls +(0,#5) and +(0,-#5) .. (#3.south); }
\newcommand{\agdeplineupupbold}[4]{ \draw[->,style=agdepstyle,line width=2pt] (#2.north) .. controls +(0,1) and +(0,1) .. (#3.north); }
\newcommand{\agdeplinedowndown}[5]{ \draw[->,style=agdepstyle] (#2.south) .. controls +(0,-#5) and +(0,-#5) .. (#3.south); }
\newcommand{\agidplinedown}[5]{ \draw[->,style=agidpstyle] (#2.south) .. controls +(0,-#5) and +(0,-#5) .. (#3.south); }
\newcommand{\agdeplinedowndownbold}[4]{ \draw[->,style=agdepstyle,line width=2pt] (#2.south) .. controls +(0,-1) and +(0,-1) .. (#3.south); }
\newcommand{\agdeplinedowndowndashed}[5]{ \draw[->,style=agidpstyle, dashed] (#2.south) .. controls +(0,-#5) and +(0,-#5) .. (#3.south); }
\newcommand{\agdeplinedownup}[4]{ \draw[->,style=agdepstyle] (#2.south) .. controls +(0,-1) and +(0,1) .. (#3.north); }
\newcommand{\agpicture}[1]{ \begin{tikzpicture}\input{#1.tikz}\end{tikzpicture} }

\newcommand{\mypic}[2]{
 \begin{figure}[ht]
  \centering
    \agpicture{diagrams/#1}
    \caption{#2}
  \label{fig:#1}
 \end{figure}
}

\newcommand{\mypicstar}[2]{
 \begin{figure*}[t]
  \centering
    \agpicture{diagrams/#1}
    \caption{#2}
  \label{fig:#1}
 \end{figure*}
}

\newcommand{\mydoublepic}[6]{
 \begin{figure*}[t]
  \centering
    \begin{minipage}[a]{#3\textwidth}
        \agpicture{diagrams/#2}
    \end{minipage}
    \begin{minipage}[a]{#5\textwidth}
        \agpicture{diagrams/#4}
    \end{minipage}
    \caption{#6}
  \label{fig:#1}
 \end{figure*}
}

\newcounter{algoctr}
\setcounter{algoctr}{-1}
\newcounter{cfctr}
\setcounter{cfctr}{-1}
\newcounter{defctr}

### 2014-10-07 paper/loags.tex
Required latex files (including a messy format.tex)

An \emph{Attribute Grammar} is a context-free grammar with inherited
and synthesized attributes assigned to every non-terminal
and a graph that represents dependencies between those attributes
at the production level.

\begin{definition}{Attribute Grammar}

An Attribute Grammar (AG) is a triple $\langle G,A,D\rangle$, where:
\begin{itemize}
 \item Context-free grammar $G = \langle N,T,P,S\rangle$ contains a set of
    non-terminals $N$, a set of terminals $T$, a set of production
    rules $P$ and a start symbol $S$. Every $p\in P$ is of the form
    $X_{p,0}\rightarrow X_{p,1}\ldots X_{p,\mid p\mid}$, with $lhs(p) = X_{p,0}$
    and $rhs(p) = X_{p,1},\ldots ,X_{p,\mid p\mid}$, where each $X_{p,i}\in
    \{lhs(p)\}\cup rhs(p)$ is called a field of $p$ and an occurrence of
    some non-terminal $X\in N$, i.e. $\exis{X\in N}{X_{p,i} = X}$.
 \item A set $A(X) = A_{inh}(X) \cup A_{syn}(X)$ is defined for all $X\in N$.
    From the attributes we infer the attribute occurrences gathered in 
    the set $A_P(p) = A_{in}(p) \cup A_{out}(p)$, where $A_{in}(p)$ 
    and $A_{out}(p)$ are the input and output occurrences of $p$ respectively.
    \begin{equation*}
     \begin{aligned}
        A_{in}(p) = &\set{X_{p,0} \cdot a}{X \cdot a \in A_{inh}(lhs(p))}\\
        \cup &\set{X_{p,i} \cdot a}{X_{p,i} \in rhs(p) 
                                \sand X = X_{p,i}
                                \sand X \cdot a \in A_{syn}(X)) }
     \end{aligned}
    \end{equation*}
    \begin{equation*}
     \begin{aligned}
        A_{out}(p) = &\set{X_{p,0} \cdot a}{X \cdot a \in A_{syn}(lhs(p))}\\
        \cup &\set{X_{p,i} \cdot a}{X_{p,i} \in rhs(p) 
                                \sand X = X_{p,i}
                                \sand X \cdot a \in A_{inh}(X)) }
     \end{aligned}
    \end{equation*}
 \item A dependency graph $D(p)$ indicates that attribute $a$ is used in
    the semantic function definition of attribute $b$ when 
    $(a\rightarrow b)\in D(p)$.
\end{itemize}
\end{definition}

The definition of AGs given above is not a complete definition in the sense that
it does not contain enough information to generate executable code from it -
the actual semantic function definitions are missing, for example.
However it contains all the information we need to define the problem of 
finding a static evaluation order for all Linear Ordered Attribute Grammars and
deciding whether an AG is an LOAG.

\begin{definition}{Linear Ordered Attribute Grammars}

A Linear Ordered Attribute Grammar (LOAG) is an AG that satisfies
the following properties:
\begin{itemize}
 \item For all $X \in N$ there exists a graph $R_X(x)$ that satisfies:
    \begin{itemize}
        \item \emph{Totality}: There must be an edge between every inherited
                and synthesized pair:
                \begin{equation*}
                 \begin{aligned}
                    \forall &(i \in A_{inh}(X), s\in A_{syn}(X))\\
                        &(i\rightarrow s) \in R_X(X) \lor (s\rightarrow i)
                    \in R_X(X)
                 \end{aligned}
                \end{equation*}
    \end{itemize}
 \item For all $p\in P$ there exists a graph $R_P(p)$ that satisfies:
    \begin{itemize}
        \item \emph{Feasibility}: The graph must include the 
            dependencies, i.e. 
            \begin{equation*}
                D(p) \subseteq R_P(p).
            \end{equation*}
        \item \emph{Consistency}: The graph must be consistent with $R_X$, i.e.
            \begin{equation*}
             \begin{aligned}
                \forall &(X\in \{lhs(p)\}\cup rhs(p), X = X_{p,i})\\
                    (&(X\cdot a\rightarrow X\cdot b)\in R_X(X) 
                    \Rightarrow (X_{p,i}\cdot a\rightarrow X_{p,i}\cdot b))
             \end{aligned}
            \end{equation*}
        \item \emph{Orderability}: The graph $R_P(p)$ must be acyclic.
    \end{itemize}
\end{itemize}
\end{definition}

Graph $R_P$ serves the same purpose as graph $ED_P$ from Kastens and from $R_X(X)$
we can infer Kastens' interfaces\cite{kastens80}.

\subsection{Satisfiability}

Using the above definition for LOAGs we can define a Boolean Satisfiability 
Problem that determines whether an arbitrary AG is an LOAG.

Firstly, be defining a variable $x_{i,s}$ for every element of the Cartesian
product $i \times s$, with $i \in A_{inh}(X)$ and $s \in A_{syn}(X)$ for all
$X\in N$. Secondly, by saying $x_{a,b} =\top$ when $a = (X_{p,i}\cdot a')$,
$b = (X_{p,i}\cdot b')$ and $(X_{p,i}\cdot a'\rightarrow X_{p,i}\cdot b')\in D_P(p)$
(\emph{feasibility}).
Thirdly, by assigning every variable $\top$ or $\bot$ (\emph{totality}) under 
the constraints that the graph $R_P$ the assignments imply is cycle free
(\emph{orderability}).
Note that \emph{consistency} is guaranteed by making sure that a variable
represents the edge between attributes (non-terminal level) as well as the edges 
between all the occurrences of that edge (production level).

%\begin{definition}{Boolean linear orderedness}
%\label{def:booleanlorderedness}
%Define a variable $x_{i,s}$ for every pair in $A_inh(X) \times A_syn(X)$ for all $X\in N$.
%An AG is a member of the class LOAG if we there is an assignment $x_{i,s} = \top$ 
%or $x_{i,s} = \bottom$ for every $x_{i,s}$, such that 
%\end{definition}

### 2014-10-07 paper/paper.tex
Required latex files (including a messy format.tex)
Add abstract and remove pdf from repo

@@
 \documentclass{llncs}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{tikz}
+\input{./formats.tex}
 
@@
 \begin{abstract}
-A nice abstract here.
-\keywords{Attribute Grammars, SAT-Solving}
+Attribute grammars are a formalism for describing tree computations (catamorphisms), and are therefore suitable for implementing the semantics of programming languages.
+A well-known subclass is the class of linear ordered attribute grammars, for which a linear evaluation order can be found statically.
+The problem of finding such an order is however known to to be NP-hard.
+
+In this paper we encode linear ordered attribute grammar scheduling as a SAT-problem.
+We encode cycle-freeness of the dependency graphs efficiently by making the graphs chordal, which is a new approach for encoding such graph property in SAT.
+We show that using Minisat as an external SAT-solver the scheduling algorithm runs much faster than existing scheduling algorithms on real-world examples.
+Furthermore, this approach allows fine-grained control over the resulting schedule, thereby enabling new attribute grammar optimizations.
+
+\keywords{Attribute Grammars, static analysis, SAT-solving}
 \end{abstract}
@@
 
+\section{Linear Ordered Attribute Grammars}
+\input{./loags.tex}
+
+\bibliographystyle{splncs}
+\bibliography{biblio}
+
 \end{document}

### 2014-10-09 paper/paper.tex
Fix typo in abstract, add Atze to authors and fix affiliations and title

@@
 \usepackage{amssymb}
+\usepackage[T1]{fontenc}
 \usepackage{tikz}
@@
 
-\author{L. Thomas van Binsbergen\inst{1} \and Jeroen Bransen\inst{2} \and Koen Claessen\inst{3}}
-\institute{todo, \email{todo@todo.co.uk}
-\and Utrecht University, Utrecht, The Netherlands, \email{J.Bransen@uu.nl}
-\and todo, \email{koen@chalmers.se}}
-\title{SAT-solving Linear Ordered Attribute Grammars}
+\title{Linear Ordered Attribute Grammar scheduling using SAT-solving}
+\author{Jeroen Bransen\inst{1}  \and L. Thomas van Binsbergen\inst{2,1} \and Koen Claessen\inst{3} \and Atze Dijkstra\inst{1}}
+\institute{Utrecht University, Utrecht, The Netherlands, \email{\{J.Bransen,atze\}@uu.nl}
+\and Royal Holloway, University of London, London, UK, \email{Thomas.VanBinsbergen.2014@live.rhul.ac.uk}
+\and Chalmers University of Technology, Gothenburg, Sweden, \email{koen@chalmers.se}}
 
@@
 A well-known subclass is the class of linear ordered attribute grammars, for which a linear evaluation order can be found statically.
-The problem of finding such an order is however known to to be NP-hard.
+The problem of finding such an order is however known to be NP-hard.
 

### 2014-10-10 paper/paper.tex
Updated abstract by Koen

@@
 \begin{abstract}
-Attribute grammars are a formalism for describing tree computations (catamorphisms), and are therefore suitable for implementing the semantics of programming languages.
-A well-known subclass is the class of linear ordered attribute grammars, for which a linear evaluation order can be found statically.
-The problem of finding such an order is however known to be NP-hard.
+Many computations over trees can be expressed using attribute
+grammars. Compilers for attribute grammars need to find an evaluation
+order (or schedule) in order to generate efficient code. For the class
+of linear ordered attribute grammars such a schedule can be found
+statically, but this problem is known to be NP-hard.
 
-In this paper we encode linear ordered attribute grammar scheduling as a SAT-problem.
-We encode cycle-freeness of the dependency graphs efficiently by making the graphs chordal, which is a new approach for encoding such graph property in SAT.
-We show that using Minisat as an external SAT-solver the scheduling algorithm runs much faster than existing scheduling algorithms on real-world examples.
-Furthermore, this approach allows fine-grained control over the resulting schedule, thereby enabling new attribute grammar optimizations.
+In this paper, we show how to encode linear ordered attribute grammar
+scheduling as a SAT-problem. Cycle-freeness of the dependency graphs
+is expressed by making the graphs chordal, which is a new approach for
+encoding such graph property in SAT.
+
+There are two main advantages to using a SAT-solver for scheduling:
+(1) the scheduling algorithm runs much faster than existing scheduling
+algorithms on real-world examples, and (2) by adding extra constraints
+we obtain fine-grained control over the resulting schedule, thereby
+enabling new scheduling optimizations.
 

### 2014-10-12 paper/paper.tex
Een aantal plaatjes met begeleidende tekst en commentaar.

@@
 \usepackage{tikz}
+\usetikzlibrary{arrows, automata}
 \input{./formats.tex}
@@
 \institute{Utrecht University, Utrecht, The Netherlands, \email{\{J.Bransen,atze\}@uu.nl}
-\and Royal Holloway, University of London, London, UK, \email{Thomas.VanBinsbergen.2014@live.rhul.ac.uk}
+\and Royal Holloway, University of London, Egham, UK, \email{Thomas.VanBinsbergen.2014@live.rhul.ac.uk}
 \and Chalmers University of Technology, Gothenburg, Sweden, \email{koen@chalmers.se}}
+
+\newif\iffinal\finalfalse
+\newcommand{\REM}[3]{\iffinal\else\textcolor{#2}{[#1: #3]}\fi}
+\newcommand{\Atze}[1]{\REM{Atze}{red}{#1}}
+\newcommand{\Koen}[1]{\REM{Koen}{red}{#1}}
+\newcommand{\Jeroen}[1]{\REM{Jeroen}{red}{#1}}
+\newcommand{\Thomas}[1]{\REM{Thomas}{red}{#1}}
 
@@
 is expressed by making the graphs chordal, which is a new approach for
-encoding such graph property in SAT.
+encoding such graph properties in SAT.
 
@@
 (1) the scheduling algorithm runs much faster than existing scheduling
-algorithms on real-world examples, and (2) by adding extra constraints
+algorithms on real-world examples\Thomas{Waarmee vergelijken we? 
+Kastens algoritme is niet voor LOAGs, het voordeel dat onze aanpak geen 
+augmenting dependencies nodig heeft zou dan in de vergelijking genoemd moeten 
+worden}, and (2) by adding extra constraints
 we obtain fine-grained control over the resulting schedule, thereby
@@
 
+\section{Attribute Grammar}
+An attribute grammars consists of three parts: a context-free grammar 
+describing abstract syntax, a set of inherited and synthesized attributes
+assigned to every non-terminal of the grammar and a set of semantic function
+definitions. A production rule of a context-free grammar has one non-terminal
+as its left-hand side and any number of terminals and non-terminals as its
+right-hand side. The non-terminals occurring in production rules are considered
+to be non-terminal \emph{occurrences}. A non-terminal occurrence of 
+non-terminal $X$ automatically obtains the attributes of $X$ as attribute 
+occurrences. The semantic function definitions describe how some attribute 
+occurrences of a production rule are calculated in terms of other attribute
+occurrences and terminal symbols of that rule. For our purposes we can safely
+ignore any terminal symbols. Figure \ref{fig:productions} shows
+examples of production rules displayed as diagrams.
+
+\mypic{productions}{Diagrams showing productions, $P1: (lhs:X) \rightarrow 
+\alpha$, $P2: (lhs:Y) \rightarrow \alpha (x:X) \beta$, $P3: (lhs:Z)
+\rightarrow \alpha (y:Y) \beta (x:X) \gamma$, where $\alpha$, $\beta$,
+$\gamma$ are arbitrary terminal symbols.}
+
+All non-terminal occurrences are named, e.g. $(x:X)$ is an occurrence of 
+non-terminal $X$ named $x$. Synthesized attributes are shown to the right
+and inherited attributes are shown to the left of their respective
+non-terminal occurrences. The gray attribute occurrences (inherited in the
+left-hand side and synthesized in the right-hand side) are input occurrences,
+the others are output occurrences.
+
+The input occurrences of one production are the output occurrences of another.
+Production rules can be `pasted' together to form a tree\cite{knuth68}. 
+This task is performed by a parser forming a derivation 
+tree\Thomas{Ik stop hier maar even met typen om 
+niet te veel uit te wijden, maar eigenlijk zou ik hier nog willen vertellen
+dat: 1) Input occurrences van buitenaf worden aangeleverd, 2) dat om dit
+te kunnen doen er wel definities voor output occurrences nodig zijn, 
+3) dat de pijlen in het figuur (omgekeerde) dependencies weergeven.}.
+
 \section{Linear Ordered Attribute Grammars}
+A linear ordered attribute grammar is an attribute grammar for which we can 
+find a evaluation order statically\Thomas{Hier zouden we misschien een 
+voorbeeld van een parse tree willen laten zien met een interessante 
+evaluation order, helaas zijn dat soort plaatjes veel te groot}. 
+That is, based on only the information of
+non-terminals and productions from the grammar, we can ensure that an 
+evaluation order exists for any valid sentence of the grammar.
+We show a static evaluation order exists by constructing evaluation orders 
+for all production rules such that they satisfy certain properties.
+Most importantly: when pasting together production rules to form a 
+derivation tree, the evaluation orders for those trees can not contradict 
+each other. Figure \ref{fig:contradiction} shows what can go wrong.
+
+\mypic{contradiction}{The evaluation orders depicted by the arrows
+of productions $P4$, $P5$ and $P6$ contradict each other: they can
+not be combined without forming a cycle.}
+
+Any non-terminal occurrence of $X$ can be replaced by any
+of the productions of $X$. Therefore, when constructing an evaluation
+order for a production in which $X$ occurs in the right-hand side,
+all the evaluation orders constructed for productions in which 
+$X$ occurs in the left-hand side need to be taken account.
+Changing the evaluation order of one production rule is likely to demand 
+a change in the evaluation order of some other production. This is 
+what makes LOAG scheduling NP-hard.
+
+In a production graph we can represent the evaluation order of other graphs
+using induced dependencies (graph $ID_P$ from Kastens) by adding an 
+arrow $a\rightarrow b$ to a non-terminal $X$ if $X$ occurs in some other
+production in which there is a path between its $a$ and $b$.
+Figure \ref{fig:idp} shows the induced dependency graph of production
+$P4$.
+
+\mypic{idp}{The gray edges are induced dependency, showing that they
+evaluation orders of Figure \ref{fig:contradiction} are not compatible.}
+
+The induced dependencies will be the same for every occurrence of every
+non-terminal and are gathered in induced dependency graphs for 
+non-terminals (Kastens $ID_S$).
+
+We can now look at the problem from two directions: 1) we change the
+evaluation order of a production and update graph $ID_S$ which might
+then affect other production rules and update until all evaluation
+orders are compatible\Thomas{AOAG} or 2) we make the decision at non-terminal 
+level and decide an order for evaluating the attributes of a non-terminal 
+such that an order at production level is still 
+possible\Thomas{LOAG, De intuitie achter dit idee zijn de visit-sequences
+en intra-visit dependencies, maar die willen we niet noemen}.
+
+\section{Chordal Graphs}
+The chords of a chordal graph can be seen as representing the assignment
+to the variables `behind' that chord. If there is a chord $a\leftrightarrow b$
+then an assignment $a\leftrightarrow^\rightarrow b$ implies that there is 
+no path $b\longrightarrow a$ assuming cycle-freeness. This intuition allows
+us to see why all assignments that lead to no cycles of length 3 lead to 
+no cycles at all in a chordal graph: every cycle of length $> 3$ is 
+split by a chord, so one side of the cycles assignments is represented by
+the chord in the other.
+
+\mypic{simplechord}{There can not exists a cycle w-n-e-s without either
+a cycle n-s-w or n-e-s.}
+
+\Thomas{In mijn scriptie gebruik ik problem graphs en assignment graphs
+om de stap van undirected naar directed graphs te maken. Later kwam ik
+echter het begrip `orientation' tegen, met hetzelfde doel.}
+
+\section{Definitions}
 \input{./loags.tex}

### 2014-10-13 paper/paper.tex
paper introduction

@@
 
+% Paper info
 \title{Linear Ordered Attribute Grammar scheduling using SAT-solving}
@@
 
+% Wat handige shortcuts
 \newif\iffinal\finalfalse
@@
 \newcommand{\Thomas}[1]{\REM{Thomas}{red}{#1}}
+
+\newcommand{\figref}[1]{Figure~\ref{#1}}
+\newcommand{\sectref}[1]{Section~\ref{#1}}
 
@@
 we obtain fine-grained control over the resulting schedule, thereby
-enabling new scheduling optimizations.
+enabling new scheduling optimisations.
 
@@
 \section{Introduction}
+Attribute Grammars \cite{knuth68} are a formalism for describing the semantics of context-free languages, thereby making them suitable for the construction of compilers.
+Examples of compilers written using attribute grammars are JastAddJ \cite{Ekman:2007}, a Java compiler, and UHC \cite{Dijkstra:2009}, a Haskell compiler.
+The UHC is also the motivation for this paper and is used as a real world test case for testing the effectiveness of the given approach.
+
+An attribute grammar essentially describes computations over trees, also known as folds or catamorphisms.
+Although the attribute grammar defines exactly \emph{what} should be computed, it does not define \emph{when} values should be computed.
+It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order.
+Such evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps.
+Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.
+
+The class of \emph{ordered attribute grammars} \cite{kastens80} is a well-known subclass of attribute grammars for which a polynomial time scheduling algorithm exists.
+However, despite what the name suggests there exist attribute grammars for which a static evaluation order can be found that do not belong to the class of ordered attribute grammars.
+In our work on compiler construction we often encountered this situation, giving rise to the need of other scheduling algorithms \cite{bransen2012}.
+We therefore look at the class of \emph{linear ordered attribute grammars} in this paper, which is the largest class of attribute grammars for which a static evaluation order can be found.
+The problem of statically finding such evaluation order is known to be NP-complete \cite{engelfriet82}.
+
+We solve the linear ordered attribute grammar scheduling problem by translating it to the Boolean satisfiability problem (SAT), one of the best-known NP-complete problems \cite{Cook:1971}.
+The SAT problem has been the subject of a lot of research and even though the worst case runtime of all known SAT-solving algorithms is exponential in the input size, many SAT-solvers work very well in practice.
+\Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating to SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
+
+The core of the scheduling problem is essentially a set of dependency graphs (directed acyclic graphs) for which a total order on the nodes should be found such that all direct dependencies, which are coming from the input grammar, are satisfied.
+However, there is a certain interplay between the different dependency graphs resulting in indirect dependencies, coming from the order that is found in one dependency graph, that need to be satisfied another dependency graph.
+This interplay of dependencies, which is explained later in detail, is what makes the problem so hard.
+
+To encode this problem in SAT we represent each edge as a Boolean variable, with its value indicating the direction of the edge.
+For the direct dependencies the value is already set, but for the rest of the variables the SAT-solver may choose the direction of the edge.
+To ensure cycle-freeness this requires us to encode transitivity with SAT-constraints, which in the straight-forward solution leads to a number of extra constraints cubing in the number of variables.
+To avoid that problem we make our graphs \emph{chordal} \cite{Dirac:1961}, which means that for every cycle of size $> 3$, there must exist an edge between two non-adjacent nodes in the cycle.
+In other words, if there exists a cycle in the graph, there must also exist a cycle of at most three nodes.
+This allows us to encode cycle-freeness much more efficiently by only disallowing cycles of length three.
+
+Apart from the fact that this translation to SAT helps in efficiently (in practice) solving the scheduling problem, there also is another benefit: it is now possible to encode extra constraints on the resulting schedule.
+We show two \Jeroen{de 2e moet nog uitgewerkt worden} scheduling optimisations that are interesting from an attribute grammar perspective, for which the optimal schedule can be found efficiently by expressing the optimisation in the SAT problem.
+
+This paper therefore makes the following main contributions:
+\begin{itemize}
+\item We show how to encode the problem of scheduling linear ordered attribute grammars as a SAT-problem
+\item We show that chordal graphs can be used to encode cycle-freeness in SAT-problems involving graph encoding
+\item We show how certain attribute grammar optimisations can be encode in the SAT-problem
+\end{itemize}
+Furthermore, we have implement the described techniques in the UUAGC\footnote{\url{http://www.cs.uu.nl/wiki/HUT/AttributeGrammarSystem}} \cite{combinator-languages} and show that the technique works well in practice for the UHC.
+
+The outline of the paper is as follows.
+We start in \sectref{sect:loag} by explaining linear ordered attribute grammars and the difficulties in scheduling in more detail.
+We then describe the translation to SAT in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}.
+In \sectref{sect:optimisations} we describe the two attribute grammar optimisations and finally in \sectref{sect:conclusion} we discuss some problems and conclude.
+
+\Jeroen{todo in intro: cite Minisat?, cite Binsbergen:2015 (PEPM)}
+
+\section{Linear Ordered Attribute Grammars} \label{sect:loag}
+\Jeroen{todo: Betere introduction LOAGs, plaatjes om scheduling toe te lichten, dependency graphs utleggen}
+
+\section{Translation to SAT} \label{sect:translation}
+\Jeroen{todo: vertaling, chordal graphs uitleggen}
+
+\section{Empirical results} \label{sect:results}
+\Jeroen{todo: getallen over UHC laten zien, het werkt echt}
+
+\section{Optimisations} \label{sect:optimisations}
+\Jeroen{todo: laten zien dat we totaal aantal visits kunnen minimaliseren en dat we sommige attributen vroeg kunnen schedulen, en intuitie geven waarom dit nuttig is (meer hebben we ook nog niet)}
+
+\section{Conclusion and discussion} \label{sect:conclusion}
+\Jeroen{todo: het werkt, misschien is andere representatie beter, en het is onvoorspelbaarder dus misschien duren sommige ags ineens heel lang om te schedulen}
+
+% -----------------------------------
 

### 2014-10-14 paper/formats.tex
Een aantal plaatjes met begeleidende tekst en commentaar.
AG section

@@
 \tikzstyle{agchldstyle}=[shape=circle,fill=red!70,minimum size=\agrelsize]
-\tikzstyle{agattrstyle}=[shape=rectangle,draw,minimum size=\agrelsize,]
+\tikzstyle{agattrstyle}=[shape=rectangle,draw,minimum size=\agrelsize]
+\tikzstyle{inputstyle}=[shape=rectangle,fill=light-gray,draw,minimum size=\agrelsize]
+\tikzstyle{outputstyle}=[shape=rectangle,draw,minimum size=\agrelsize]
 \tikzstyle{agdepstyle}=[black,>=triangle 60]
 \tikzstyle{agidpstyle}=[gray,>=triangle 60]
-\tikzstyle{inputstyle}=[shape=rectangle,fill=light-gray,draw,minimum size=\agrelsize]
-\tikzstyle{outputstyle}=[shape=rectangle,draw,minimum size=\agrelsize]
 \newcommand{\agcoordpary}{6}   %% y-coordinate of parent attributes
@@
 \newcommand{\agchldattr}[6]{ \draw (#4 \agrelsize,\agcoordchldy \agrelsize) node [style=#6] (#3) {}; \draw (#3.north) node [anchor=north] {\tiny{#2}}; }
+\newcommand{\aglocattr}[5]{ \draw (#4 \agrelsize,\agcoordlocy \agrelsize) node [style=agattrstyle] (#3) {}; \draw (#3.south) node [anchor=north] {\tiny{ #2 }}; }
 \newcommand{\agchldnonterm}[5]{ \draw (#4 \agrelsize,\agcoordchldy \agrelsize) node [style=agnontermstyle] (#3) {}; \draw (#3.south) node [anchor=north] {\footnotesize{ #2:#5 }}; }
@@
 \newcommand{\agdeplineupdown}[4]{ \draw[->,style=agdepstyle] (#2.north) .. controls +(0,1) and +(0,-1) .. (#3.south); }
-\newcommand{\agdeplineupup}[5]{ \draw[->,style=agdepstyle] (#2.north) .. controls +(0,#5) and +(0,#5) .. (#3.north); }
+\newcommand{\agdeplineupup}[4]{ \draw[->,style=agdepstyle] (#2.north) .. controls +(0,1) and +(0,1) .. (#3.north); }
+\newcommand{\agdeplineupupctrl}[5]{ \draw[->,style=agdepstyle] (#2.north) .. controls +(0,#5) and +(0,#5) .. (#3.north); }
 \newcommand{\agidplineup}[5]{ \draw[->,style=agidpstyle] (#2.north) .. controls +(0,#5) and +(0,#5) .. (#3.north); }
-\newcommand{\agidplineupdown}[5]{ \draw[->,style=agidpstyle] (#2.north) .. controls +(0,#5) and +(0,-#5) .. (#3.south); }
 \newcommand{\agdeplineupupbold}[4]{ \draw[->,style=agdepstyle,line width=2pt] (#2.north) .. controls +(0,1) and +(0,1) .. (#3.north); }
-\newcommand{\agdeplinedowndown}[5]{ \draw[->,style=agdepstyle] (#2.south) .. controls +(0,-#5) and +(0,-#5) .. (#3.south); }
+\newcommand{\agdeplinedowndown}[4]{ \draw[->,style=agdepstyle] (#2.south) .. controls +(0,-1) and +(0,-1) .. (#3.south); }
+\newcommand{\agdeplinedowndownctrl}[5]{ \draw[->,style=agdepstyle] (#2.south) .. controls +(0,-#5) and +(0,-#5) .. (#3.south); }
 \newcommand{\agidplinedown}[5]{ \draw[->,style=agidpstyle] (#2.south) .. controls +(0,-#5) and +(0,-#5) .. (#3.south); }
@@
 \newcommand{\agdeplinedownup}[4]{ \draw[->,style=agdepstyle] (#2.south) .. controls +(0,-1) and +(0,1) .. (#3.north); }
+\newcommand{\agvarlinedownup}[6]{ \draw[<->,style=agdepstyle] (#2.south) .. controls +(0,-1) and +(0,1) .. node[#5] {#6} (#3.north); }
+\newcommand{\agvarlinedowndown}[6]{ \draw[->,style=agidpstyle] (#2.south) .. controls +(0,-1) and +(0,-1) .. node[#5] {#6} (#3.south); }
+\newcommand{\agvarlineupup}[6]{ \draw[->,style=agidpstyle] (#2.north) .. controls +(0,1) and +(0,1) .. node[#5] {#6} (#3.north); }
 \newcommand{\agpicture}[1]{ \begin{tikzpicture}\input{#1.tikz}\end{tikzpicture} }
 
+\newcommand{\exprrelsize}{0.20cm}
+\newcommand{\exprdist}{0.20cm}
+\tikzstyle{litnode}=[circle,draw,minimum size=\exprrelsize]
+\tikzstyle{astline}=[]
+\tikzstyle{synnode}=[rectangle,draw,black,minimum size=\exprrelsize]
+\tikzstyle{inhnode}=[rectangle,draw,black,minimum size=\exprrelsize]
+\tikzstyle{synline}=[gray,->, thick]
+\tikzstyle{chord}=[black,<->, thick]
+\tikzstyle{inhline}=[gray,->, thick]
+
 \newcommand{\mypic}[2]{
- \begin{figure}[ht]
+ \begin{figure}
   \centering
@@
 
-\newcommand{\mydoublepic}[6]{
- \begin{figure*}[t]
+\newcommand{\mydoublepic}[5]{
+ \begin{figure}
   \centering
-    \begin{minipage}[a]{#3\textwidth}
-        \agpicture{diagrams/#2}
+    \begin{minipage}[a]{#2\textwidth}
+        \agpicture{diagrams/#1}
     \end{minipage}
-    \begin{minipage}[a]{#5\textwidth}
-        \agpicture{diagrams/#4}
+    \begin{minipage}[a]{#4\textwidth}
+        \agpicture{diagrams/#3}
     \end{minipage}
-    \caption{#6}
+    \caption{#5}
   \label{fig:#1}
- \end{figure*}
+ \end{figure}
 }

### 2014-10-14 paper/paper.tex
AG section

@@
 The outline of the paper is as follows.
-We start in \sectref{sect:loag} by explaining linear ordered attribute grammars and the difficulties in scheduling in more detail.
+We start in \sectref{sect:ag} and \sectref{sect:loag} by explaining linear ordered attribute grammars and the difficulties in scheduling in more detail.
 We then describe the translation to SAT in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}.
@@
 
+\section{Attribute Grammars} \label{sect:ag}
+Formally an attribute grammar consists of three parts: a context-free grammar, a set of attribute definitions and a set of semantic functions.
+Instead of a context-free grammar describing the concrete syntax of the language, we feel that it is more intuitive to visualise the grammar as describing the abstract syntax tree of the language.
+We therefore say that an attribute grammar consists of a set of algebraic data types describing the abstract syntax, a set of attribute definitions and a set of semantic functions.
+The attribute definitions and semantic functions describe how values should be computed over the abstract syntax tree.
+
+In the rest of this paper we do stick to the usual attribute grammar terminology, so with a \emph{nonterminal} we mean a type, and a \emph{production} is a constructor of that type.
+Furthermore, the name \emph{lhs} (short for left-hand side) refers to the constructor itself, while the children of a production have separate names.
+As this paper is about scheduling and data dependencies, we do not explain the syntax of the semantic functions here, but it is important to remark that these functions define how the value of a certain attribute can be computed from other attributes.
+
+There are two types of attributes: \emph{inherited} attributes with values that are passed from a parent to its children, and \emph{synthesized} attributes with values that are passed from the children to their parents.
+In the pictures we draw inherited attributes on the left side of a node, and the synthesized attributes on its right side.
+
+\mydoublepic{labval_bin_flow_full}{0.55}{labval_leaf_flow_full}{0.3}{
+    Example attribute grammar for binary trees with two inherited and two synthesized attributes}
+
+In \figref{fig:labval_bin_flow_full} we show an example attribute grammar with one nonterminal \emph{Tree} which has two productions: \emph{Bin} and \emph{Leaf}.
+The \emph{Bin} production has two children named \emph{l} and \emph{r}, both of the type \emph{Tree}, and \emph{Leaf} has no children.
+The \emph{Tree} nonterminal has four attributes: two inherited attributes a1 and a2 and two synthesized attributes a3 and a4.
+
+The figure shows what is called the \emph{production dependency graphs}.
+For a production two different types of attributes can be distinguished: \emph{input} attributes, which are attributes for which the value is provided, and \emph{output} attributes, which are values that need to be defined.
+The input attributes of a production, colored gray in the picture, are the inherited attributes of the production itself, because their values are provided by the parent node, and the synthesized attributes of all its children.
+The output attributes, colored white in the picture, are the synthesized attributes of the parent and the inherited attributes of the child.
+
+Although we talk about dependency graphs in the context of scheduling, we actually draw the edges in the other direction.
+The edges in the picture denote data flow, which from an attribute grammar perspective is much more intuitive.
+For example, in the \emph{Bin} production the attribute a1 of the child \emph{r} is computed from a1 from the parent and a3 from the child \emph{l}.
+The edges are thus always directed from an input attribute to an output attribute, and the true dependency graph can be obtained by reversing all edges.
+
 \section{Linear Ordered Attribute Grammars} \label{sect:loag}
-\Jeroen{todo: Betere introduction LOAGs, plaatjes om scheduling toe te lichten, dependency graphs utleggen}
+\Jeroen{todo: vertel wat subklasse van LOAG inhoudt; nonterminal dependency graphs, en de interactie in scheduling}
 
@@
 \Jeroen{todo: het werkt, misschien is andere representatie beter, en het is onvoorspelbaarder dus misschien duren sommige ags ineens heel lang om te schedulen}
+
+\bibliographystyle{splncs}
+\bibliography{biblio}
+
+\end{document}
 

### 2014-10-14 paper/paper.tex
Cite PEPM paper in introduction

@@
 \newcommand{\sectref}[1]{Section~\ref{#1}}
+
+% Handle citing "Van Binsbergen" correctly...
+\DeclareRobustCommand{\VAN}[3]{#2}
 
@@
 However, despite what the name suggests there exist attribute grammars for which a static evaluation order can be found that do not belong to the class of ordered attribute grammars.
-In our work on compiler construction we often encountered this situation, giving rise to the need of other scheduling algorithms \cite{bransen2012}.
+In our work on compiler construction we often encountered this situation, giving rise to the need of other scheduling algorithms \cite{bransen2012,Binsbergen:2015}.
 We therefore look at the class of \emph{linear ordered attribute grammars} in this paper, which is the largest class of attribute grammars for which a static evaluation order can be found.
@@
 
-\Jeroen{todo in intro: cite Minisat?, cite Binsbergen:2015 (PEPM)}
+\Jeroen{todo in intro: cite Minisat?}
 
@@
 
+\DeclareRobustCommand{\VAN}[3]{#3}
 \bibliographystyle{splncs}

### 2014-10-14 paper/paper.tex
LOAG section

@@
 \section{Linear Ordered Attribute Grammars} \label{sect:loag}
-\Jeroen{todo: vertel wat subklasse van LOAG inhoudt; nonterminal dependency graphs, en de interactie in scheduling}
+Given an attribute grammar definition the attribute grammar compiler should produce a runtime evaluator that takes an abstract syntax tree as input and computes all attribute values.
+There are two approaches for doing the scheduling: at runtime (dynamic) or at compile time (static).
+
+For dynamic scheduling one could rely on existing machinery for lazy evaluation, for example in Haskell, which is the approach taken by \cite{saraiva99}.
+There the attribute grammar definitions are translated to lazy Haskell functions in a straightforward way by producing functions that take the inherited attributes as argument and return the synthesized attributes.
+Whenever an inherited attribute (indirectly) depends on the value of a synthesized attribute, this means the function has a cyclic definition, which is no problem for languages with lazy semantics, and can actually result in effient code \cite{Bird:1984}.
+
+There are two problems with the dynamic scheduling.
+First of all whenever the attribute grammar contains a true cyclic attribute definition, the evaluator enters an infinite loop at runtime.
+However, these cycles could have been detected at compile time!
+Furthermore, the code needs lazy functions while with the static scheduling strict evaluators can be generated, leading to more effient code in practice.
+
+For static scheduling efficient evaluators can be generated, but the problem is that static scheduling is hard.
+In \cite{kastens80} a polynomial time algorithm is given for static scheduling of the class of ordered attribute grammars, but unfortunately we often encounter attribute grammars outside of that class.
+All our attribute grammars do however fall in the class of linear ordered attribute grammars, which is the largest class of attribute grammars for which a static schedule can be found.
+Although the scheduling is NP-complete, we have implemented a backtracking algorithm in earlier work \cite{Binsbergen:2015} that is feasible in practice.
+However, in this paper we show that we can do better.
+
+Conceptually, static scheduling of linear ordered attribute grammars is not very complex.
+For each nonterminal a total order should be found on its attributes, such that all production dependency graphs all cycle free.
+However, because a nonterminal can have multiple productions and a production can have children of different types, choices made on the order of attributes of one nonterminal can influence the order of attributes in another nonterminal.
+
+\mypic{labval_bin_flow_induced}{Production dependency graph for Bin for the order a2 $\rightarrow$ a4 $\rightarrow$ a1 $\rightarrow$ a3}
+
+In \figref{fig:labval_bin_flow_induced} we show the production dependency graph of the nonterminal \emph{Bin} for a complete order on the attributes.
+In this case there is only a single nonterminal so both the parent and the child nodes have the same order.
+Because this dependency graph is still cycle free after adding the edges for the complete order and the same holds for the \emph{Leaf} production, the order a2 $\rightarrow$ a4 $\rightarrow$ a1 $\rightarrow$ a3 is a valid schedule for the nonterminal \emph{Tree}.
 

### 2014-10-14 paper/paper.tex
Translation section

@@
 
+In order to encode this we also construct a \emph{nonterminal dependency graph}.
+This graph contains all attributes that are defined for the corresponding nonterminal and the edges in this graph must define a total order on the attributes.
+Furthermore, the edges in the nonterminal dependency graph should agree with the (indirect) dependencies from the production dependency graphs such that no cycles exist.
+
 \mypic{labval_bin_flow_induced}{Production dependency graph for Bin for the order a2 $\rightarrow$ a4 $\rightarrow$ a1 $\rightarrow$ a3}
 
-In \figref{fig:labval_bin_flow_induced} we show the production dependency graph of the nonterminal \emph{Bin} for a complete order on the attributes.
-In this case there is only a single nonterminal so both the parent and the child nodes have the same order.
+In \figref{fig:labval_bin_flow_induced} we show the production dependency graph of the nonterminal \emph{Bin} with a complete order on the attributes.
+In this case there is only a single nonterminal so both the parent and the child nodes have the same attribute order, made explicit by extra edges from the nonterminal dependency graph.
 Because this dependency graph is still cycle free after adding the edges for the complete order and the same holds for the \emph{Leaf} production, the order a2 $\rightarrow$ a4 $\rightarrow$ a1 $\rightarrow$ a3 is a valid schedule for the nonterminal \emph{Tree}.
@@
 \section{Translation to SAT} \label{sect:translation}
-\Jeroen{todo: vertaling, chordal graphs uitleggen}
+To represent the scheduling problem as a Boolean formula we introduce a variable for each edge, indicating the direction of the edge.
+The direct dependencies coming from the source code are constants, but for the rest of the edges the SAT solver can decide on the direction.
+However, the encoding has been chosen in such way that a valid assigment of the variables corresponds to a valid schedule.
+
+Our algorithm has the following steps:
+\begin{enumerate}
+\item Construct a nonterminal dependency graph for each nonterminal
+\item Construct a production dependency graph for each production
+\item Make all graphs chordal
+\item Introduce a SAT variable for each edge in any of the graphs, where edges from a production dependency graph and the corresponding nonterminal dependency graph are shared, so are represented by the same variable
+\item Set the value of all variables corresponding to direct dependencies
+\item Exclude all cycles of length three
+\item Optionally add extra constraints for optimisations
+\end{enumerate}
+The first two steps have been explained in the previous sections. Step 3 is explained below, and step 5 and 6 follow from that. Finally step 7 is explained in \sectref{sect:optimisations}.
+
+\subsection{Chordal graphs}
+A chordal graph is an undirected graph in which each cycle of length $> 3$ contains a \emph{chord}.
+A chord is an edge between two nodes in the cycle that are not adjacent.
+The result of this is that each cycle of length $> 3$ is split up into two smaller cycles, so if a chordal graph contains a cycle it must also contain a cycle of size three.
+Chordal graphs are therefore sometimes also referred to as \emph{triangulated graphs}.
+
+\mypic{simplechord}{There can not exists a cycle w-n-e-s without either
+a cycle n-s-w or n-e-s.}
+
+In our case the graphs are directed, but we can still apply the same trick!
+In \figref{fig:simplechord} we illustrate how to use chordal graphs to exclude cycles.
+If there exists a cycle of length four, then it is not possible to choose a direction for the edge between \emph{n} and \emph{s} without introducing a cycle of length three.
+Hence, if we make our graph chordal by adding edges which may have an arbitrary direction and explicitly exclude all cycles of length three, we ensure that no cycles can exist at all.
+
+\Jeroen{todo: say something about heuristics for making the graph chordal and how many triangles there are}
+
+\subsection{Finding the schedule}
+When the constructed Boolean formula is given to the SAT-solver, the result is either that the formula is not satisfiable, meaning that no schedule has been found, of satisfiable, meaning that there is a schedule.
+It is not hard to see that the formula is satisfiable if and only if there exists a valid schedule for the given attribute grammar definition, and in this paper we give no formal proof of this claim.
+
+In the case where the formula is satisfiable, we obviously want to find the result.
+From the SAT solver we can ask for the truth value of each variable in the solution, so when our algorithm keeps the connection between edges and variables we can complete our directed graph and trivially find the complete order for all attributes from that.
+The constraints guarantee that this graph contains no cycles.
 

### 2014-10-14 paper/paper.tex
Add Minisat reference and two subtitles to introduction

@@
 It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order.
-Such evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps.
+Such an evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps.
 Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.
@@
 
+\subsection{Summary}
 We solve the linear ordered attribute grammar scheduling problem by translating it to the Boolean satisfiability problem (SAT), one of the best-known NP-complete problems \cite{Cook:1971}.
@@
 \Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating to SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
+In our implementation we have used MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.
 
@@
 However, there is a certain interplay between the different dependency graphs resulting in indirect dependencies, coming from the order that is found in one dependency graph, that need to be satisfied another dependency graph.
-This interplay of dependencies, which is explained later in detail, is what makes the problem so hard.
+This interplay of dependencies, which is explained later in detail, is what makes the problem hard.
 
@@
 
-This paper therefore makes the following main contributions:
+\subsection{Overview}
+In this paper we make the following main contributions:
 \begin{itemize}
@@
 In \sectref{sect:optimisations} we describe the two attribute grammar optimisations and finally in \sectref{sect:conclusion} we discuss some problems and conclude.
-
-\Jeroen{todo in intro: cite Minisat?}
 
@@
 Although the scheduling is NP-complete, we have implemented a backtracking algorithm in earlier work \cite{Binsbergen:2015} that is feasible in practice.
-However, in this paper we show that we can do better.
+However, in this paper we show that we can do better\Thomas{Faster and allowing for the different optimisations}.
 
@@
 In order to encode this we also construct a \emph{nonterminal dependency graph}.
-This graph contains all attributes that are defined for the corresponding nonterminal and the edges in this graph must define a total order on the attributes.
+This graph contains all attributes that are defined for the corresponding nonterminal and the edges in this graph must define a total order\Thomas{Actually it can be partial, which is also what we obtain using a variable for every inherited x synthesized pair.
+For example, an interface with only 1 visit (and multiple inherited or synthesized attributes) defines a partial order since we don't care about the order between inherited/synthesized attributes in the same visit.
+The visit-sequences that are calculated in the end do imply a total order however.} on the attributes.
 Furthermore, the edges in the nonterminal dependency graph should agree with the (indirect) dependencies from the production dependency graphs such that no cycles exist.
@@
 \end{enumerate}
-The first two steps have been explained in the previous sections. Step 3 is explained below, and step 5 and 6 follow from that. Finally step 7 is explained in \sectref{sect:optimisations}.
+The first two steps have been explained in the previous sections.
+Step 3 is explained below\Thomas{Step 4 is missing}, and step 5 and 6 follow from that.
+Finally step 7 is explained in \sectref{sect:optimisations}.
 

### 2014-10-15 paper/paper.tex
Added some comments

@@
 
-\subsection{Summary}
 We solve the linear ordered attribute grammar scheduling problem by translating it to the Boolean satisfiability problem (SAT), one of the best-known NP-complete problems \cite{Cook:1971}.
@@
 \Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating to SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
-In our implementation we have used MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.
 
@@
 
-\subsection{Overview}
-In this paper we make the following main contributions:
+This paper therefore makes the following main contributions:
 \begin{itemize}
@@
 In \sectref{sect:optimisations} we describe the two attribute grammar optimisations and finally in \sectref{sect:conclusion} we discuss some problems and conclude.
+
+\Jeroen{todo in intro: cite Minisat?}
 

### 2014-10-15 paper/paper.tex
Subsection on chordal graph construction

--- 2014-10-14 paper/paper.tex
@@
 The first two steps have been explained in the previous sections.
-Step 3 is explained below\Thomas{Step 4 is missing}, and step 5 and 6 follow from that.
+Step 3 is explained below, step 4 is trivial, and step 5 and 6 follow from the explanation below.
 Finally step 7 is explained in \sectref{sect:optimisations}.
@@
 
-\Jeroen{todo: say something about heuristics for making the graph chordal and how many triangles there are}
+\subsection{Chordal graph construction}
+There are several algorithms for making a graph chordal. We use an algorithm based on the following alternative definition of a chordal graph:
+
+\begin{definition}
+An undirected graph is \emph{chordal} if and only if it has a 
+perfect elimination order. A perfect elimination order is an
+ordering $v_1,\ldots,v_n$ of the vertices of $G$ such that in the graph
+$G[v_1,\ldots,v_i]$, $\univ{i}{1 \leq i \leq n}$, the vertex $v_i$ is 
+\toindex{simplicial}. A vertex $v$ is called simplicial in a graph $G$ if
+the neighbourhood of $v$ forms a connected component in $G$.
+The graph $G[v_1,\ldots,v_i]$ is the induced subgraph of $G$ containing
+only the vertices $v_i,\ldots,v_i$ and the edges between these vertices.
+\end{definition}
+
+From this definition we can construct the following algorithm for making a graph chordal:
+\begin{enumerate}
+ \item While the graph still contains vertices:
+ \begin{enumerate}
+     \item Select a vertex $v$ from the graph
+     \item For every pair $(a,b)$ of unconnected vertices in the neighbourhood
+        of $v$: 
+        \begin{enumerate}
+            \item Add the edge $(a\leftrightarrow b)$ to the graph
+        \end{enumerate}
+     \item Remove $v$, and all edges connected to $v$, from the graph.
+ \end{enumerate}
+\end{enumerate}
+
+One important open question in this algorithm is the order in which the vertices should be chosen.
+In \sectref{sect:results} we show the results for several heuristics that we have implemented and tried on the UHC.
+We would like to remark that regardless of the heuristic used, this approach always leads to much smaller SAT problems than encoding transitivity in the SAT problem for ruling out cycles.
 

### 2014-10-15 paper/paper.tex
Results section

@@
 \newcommand{\figref}[1]{Figure~\ref{#1}}
+\newcommand{\tabref}[1]{Table~\ref{#1}}
 \newcommand{\sectref}[1]{Section~\ref{#1}}
@@
 
-\subsection{Chordal graph construction}
+\subsection{Chordal graph construction} \label{sect:chordalconstruct}
 There are several algorithms for making a graph chordal. We use an algorithm based on the following alternative definition of a chordal graph:
@@
 One important open question in this algorithm is the order in which the vertices should be chosen.
-In \sectref{sect:results} we show the results for several heuristics that we have implemented and tried on the UHC.
+In \sectref{sect:heuristics} we show the results for several heuristics that we have implemented and tried on the UHC.
 We would like to remark that regardless of the heuristic used, this approach always leads to much smaller SAT problems than encoding transitivity in the SAT problem for ruling out cycles.
@@
 \section{Empirical results} \label{sect:results}
-\Jeroen{todo: getallen over UHC laten zien, het werkt echt}
+As a test case for our work we use the Utrecht Haskell Compiler (UHC) \cite{Dijkstra:2009}.
+The source code of the UHC consists of several attribute grammar definitions together with Haskell code.
+The biggest attribute grammar in the UHC, called \emph{MainAG}, consists of 30 nonterminals, 134 productions, 1332 attributes (44.4 per nonterminal) and 9766 dependencies and is the biggest attribute grammar we know of.
+
+The UHC is compiled using the Utrecht University Attribute Grammar Compiler (UUAGC) \cite{combinator-languages}, in which we have implemented our approach.
+Apart from our SAT approach there are three other scheduling algorithms that have been implemented to which we can compare our approach:
+
+\begin{itemize}
+\item \emph{Kastens}: the algorithm from \cite{kastens80} which only works for ordered attribute grammars, of which MainAG is not a member.
+However, with extra manual annotations this algorithm can generate a schedule, which is the approach used previously.
+However, as writing these annotation is hard an error-prone, this is not an optimal solution.
+\item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linear ordered attribute grammars.
+However, the scheduling is not completely static so the generated evaluator also does part of the scheduling at runtime.
+\item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm \cite{Binsbergen:2015} for linear ordered attribute grammars, based on Kastens' algorithm.
+This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper.
+\end{itemize}
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{l || c || r}
+      Algorithm   &   Manual annotations   &  Compile time MainAG \\
+      \hline
+      Kastens'                &   Y           &   25s \\
+      Kennedy-Warren  &   N           &   2s \\
+      LOAG-backtrack  &   N           &   35s \\
+      LOAG-SAT           &   N           &   10s \\
+    \end{tabular}
+  \end{center}
+  \caption{Comparison of the four scheduling algorithms}
+  \label{tab:algo-comparison}
+\end{table}
+
+In \tabref{tab:algo-comparison} we show the compilation times for the MainAG of the UHC for each of the four different algorithms.
+This time includes parsing of the attribute grammar description and code generation, which is what takes most time in the SAT approach.
+The SAT-solver takes less than a second to find a solution.
+\Jeroen{De tijd voor Kennedy-Warren kan niet kloppen, Thomas kun jij die nog eens checken?
+Op mijn machine doen alle opties er veel langer over...}
+
+\subsection{Chordal graph heuristics} \label{sect:heuristics}
+\begin{table}
+  \begin{center}
+    \begin{tabular}{c r r r }
+      Order                                   &   \#Clauses   & \#Vars       & Ratio \\
+      \hline
+      $(\mid\mathcal{D}\mid,\mid\mathcal{S}\mid,\mid\mathcal{C}\mid)$ &  21,307,812  & \quad 374,792           & \quad 57.85 \\
+      $(\mid\mathcal{D}\mid,\mid\mathcal{C}\mid,\mid\mathcal{S}\mid)$ &   8,301,557   & 220,690           & 37.62 \\
+      $(\mid\mathcal{S}\mid,\mid\mathcal{D}\mid,\mid\mathcal{C}\mid)$ &  12,477,519   & 287,151           & 43.45 \\
+      $(\mid\mathcal{S}\mid,\mid\mathcal{C}\mid,\mid\mathcal{D}\mid)$ &   8,910,379   & 241,853           & 36.84 \\
+      $(\mid\mathcal{C}\mid,\mid\mathcal{D}\mid,\mid\mathcal{S}\mid)$ &   3,004,705   & 137,277           & 21.89 \\
+      $(\mid\mathcal{C}\mid,\mid\mathcal{S}\mid,\mid\mathcal{D}\mid)$ &   3,359,910   & 156,795           & 21.43 \\
+      $(\mid\mathcal{D}\mid+\mid\mathcal{S}\mid,\mid\mathcal{C}\mid)$     &  12,424,635   & 386,323           & 32.16 \\
+      $(\mid\mathcal{D}\mid,\mid\mathcal{S}\mid+\mid\mathcal{C}\mid)$     &   8,244,600   & 219,869           & 37.50 \\
+      $(\mid\mathcal{D}\mid+\mid\mathcal{C}\mid,\mid\mathcal{S}\mid)$     &   2,930,922   & 135,654           & 21.61 \\
+      $(\mid\mathcal{S}\mid,\mid\mathcal{D}\mid+\mid\mathcal{C}\mid)$     &   8,574,307   & 236,348           & 36.28 \\
+      $(\mid\mathcal{S}\mid+\mid\mathcal{C}\mid,\mid\mathcal{D}\mid)$     &   3,480,866   & 157,089           & 22.16 \\
+      $(\mid\mathcal{C}\mid,\mid\mathcal{D}\mid+\mid\mathcal{S}\mid)$     &   3,392,930   & 157,568           & 21.53 \\
+      $(\mid\mathcal{C}\mid+\mid\mathcal{D}\mid+\mid\mathcal{S}\mid)$     &   3,424,001   & 148,724           & 23.02 \\
+      $(3*\mid\mathcal{S}\mid*(\mid\mathcal{D}\mid+\mid\mathcal{C}\mid)+(\mid\mathcal{D}\mid*\mid\mathcal{C}\mid) ^2)$     &   \underline{2,679,772}   & \underline{127,768}           & \underline{20.97} \\ \\
+    \end{tabular}
+  \end{center}
+  \caption{Table showing the number of clauses and variables required for solving the MainAG of the UHC, selecting the next vertex in the elimination order based on different ways to compare neighbourhoods.}
+  \label{tab:edgecomp}
+\end{table}
+
+As explained in \sectref{sect:chordalconstruct} we need to find an order in which to handle the vertices such that the resulting SAT problem is as small as possible.
+In \tabref{tab:edgecomp} we show the results of different heuristics for the MainAG of the UHC.
+In this table we use three different sets: $\mathcal{D}$ is the set of direct dependencies, $\mathcal{C}$ is the set of edges that are added to make the graph chordal and $\mathcal{S}$ is the set of other edges.
+For each of the sets we take only the edges in the neigbourhood of \emph{v} for comparison.
 

### 2014-10-15 paper/paper.tex
Respond to / solve Thomas' comments

@@
 (1) the scheduling algorithm runs much faster than existing scheduling
-algorithms on real-world examples\Thomas{Waarmee vergelijken we? 
-Kastens algoritme is niet voor LOAGs, het voordeel dat onze aanpak geen 
-augmenting dependencies nodig heeft zou dan in de vergelijking genoemd moeten 
-worden}, and (2) by adding extra constraints
+algorithms on real-world examples, and (2) by adding extra constraints
 we obtain fine-grained control over the resulting schedule, thereby
@@
 Although the scheduling is NP-complete, we have implemented a backtracking algorithm in earlier work \cite{Binsbergen:2015} that is feasible in practice.
-However, in this paper we show that we can do better\Thomas{Faster and allowing for the different optimisations}.
+However, in this paper we show that we can do better by creating an algorithm that is faster and allows for different optimisations on the resulting schedule.
 
@@
 For example, an interface with only 1 visit (and multiple inherited or synthesized attributes) defines a partial order since we don't care about the order between inherited/synthesized attributes in the same visit.
-The visit-sequences that are calculated in the end do imply a total order however.} on the attributes.
+The visit-sequences that are calculated in the end do imply a total order however.}\Jeroen{Yes, but it can also just be total, which is much easier to explain, and I think this detail doesn't matter here} on the attributes.
 Furthermore, the edges in the nonterminal dependency graph should agree with the (indirect) dependencies from the production dependency graphs such that no cycles exist.

### 2014-10-15 paper/paper.tex
Remove leftover stuff from paper

@@
 \end{document}
-
-% -----------------------------------
-
-\section{Attribute Grammar}
-An attribute grammars consists of three parts: a context-free grammar 
-describing abstract syntax, a set of inherited and synthesized attributes
-assigned to every non-terminal of the grammar and a set of semantic function
-definitions. A production rule of a context-free grammar has one non-terminal
-as its left-hand side and any number of terminals and non-terminals as its
-right-hand side. The non-terminals occurring in production rules are considered
-to be non-terminal \emph{occurrences}. A non-terminal occurrence of 
-non-terminal $X$ automatically obtains the attributes of $X$ as attribute 
-occurrences. The semantic function definitions describe how some attribute 
-occurrences of a production rule are calculated in terms of other attribute
-occurrences and terminal symbols of that rule. For our purposes we can safely
-ignore any terminal symbols. Figure \ref{fig:productions} shows
-examples of production rules displayed as diagrams.
-
-\mypic{productions}{Diagrams showing productions, $P1: (lhs:X) \rightarrow 
-\alpha$, $P2: (lhs:Y) \rightarrow \alpha (x:X) \beta$, $P3: (lhs:Z)
-\rightarrow \alpha (y:Y) \beta (x:X) \gamma$, where $\alpha$, $\beta$,
-$\gamma$ are arbitrary terminal symbols.}
-
-All non-terminal occurrences are named, e.g. $(x:X)$ is an occurrence of 
-non-terminal $X$ named $x$. Synthesized attributes are shown to the right
-and inherited attributes are shown to the left of their respective
-non-terminal occurrences. The gray attribute occurrences (inherited in the
-left-hand side and synthesized in the right-hand side) are input occurrences,
-the others are output occurrences.
-
-The input occurrences of one production are the output occurrences of another.
-Production rules can be `pasted' together to form a tree\cite{knuth68}. 
-This task is performed by a parser forming a derivation 
-tree\Thomas{Ik stop hier maar even met typen om 
-niet te veel uit te wijden, maar eigenlijk zou ik hier nog willen vertellen
-dat: 1) Input occurrences van buitenaf worden aangeleverd, 2) dat om dit
-te kunnen doen er wel definities voor output occurrences nodig zijn, 
-3) dat de pijlen in het figuur (omgekeerde) dependencies weergeven.}.
-
-\section{Linear Ordered Attribute Grammars}
-A linear ordered attribute grammar is an attribute grammar for which we can 
-find a evaluation order statically\Thomas{Hier zouden we misschien een 
-voorbeeld van een parse tree willen laten zien met een interessante 
-evaluation order, helaas zijn dat soort plaatjes veel te groot}. 
-That is, based on only the information of
-non-terminals and productions from the grammar, we can ensure that an 
-evaluation order exists for any valid sentence of the grammar.
-We show a static evaluation order exists by constructing evaluation orders 
-for all production rules such that they satisfy certain properties.
-Most importantly: when pasting together production rules to form a 
-derivation tree, the evaluation orders for those trees can not contradict 
-each other. Figure \ref{fig:contradiction} shows what can go wrong.
-
-\mypic{contradiction}{The evaluation orders depicted by the arrows
-of productions $P4$, $P5$ and $P6$ contradict each other: they can
-not be combined without forming a cycle.}
-
-Any non-terminal occurrence of $X$ can be replaced by any
-of the productions of $X$. Therefore, when constructing an evaluation
-order for a production in which $X$ occurs in the right-hand side,
-all the evaluation orders constructed for productions in which 
-$X$ occurs in the left-hand side need to be taken account.
-Changing the evaluation order of one production rule is likely to demand 
-a change in the evaluation order of some other production. This is 
-what makes LOAG scheduling NP-hard.
-
-In a production graph we can represent the evaluation order of other graphs
-using induced dependencies (graph $ID_P$ from Kastens) by adding an 
-arrow $a\rightarrow b$ to a non-terminal $X$ if $X$ occurs in some other
-production in which there is a path between its $a$ and $b$.
-Figure \ref{fig:idp} shows the induced dependency graph of production
-$P4$.
-
-\mypic{idp}{The gray edges are induced dependency, showing that they
-evaluation orders of Figure \ref{fig:contradiction} are not compatible.}
-
-The induced dependencies will be the same for every occurrence of every
-non-terminal and are gathered in induced dependency graphs for 
-non-terminals (Kastens $ID_S$).
-
-We can now look at the problem from two directions: 1) we change the
-evaluation order of a production and update graph $ID_S$ which might
-then affect other production rules and update until all evaluation
-orders are compatible\Thomas{AOAG} or 2) we make the decision at non-terminal 
-level and decide an order for evaluating the attributes of a non-terminal 
-such that an order at production level is still 
-possible\Thomas{LOAG, De intuitie achter dit idee zijn de visit-sequences
-en intra-visit dependencies, maar die willen we niet noemen}.
-
-\section{Chordal Graphs}
-The chords of a chordal graph can be seen as representing the assignment
-to the variables `behind' that chord. If there is a chord $a\leftrightarrow b$
-then an assignment $a\leftrightarrow^\rightarrow b$ implies that there is 
-no path $b\longrightarrow a$ assuming cycle-freeness. This intuition allows
-us to see why all assignments that lead to no cycles of length 3 lead to 
-no cycles at all in a chordal graph: every cycle of length $> 3$ is 
-split by a chord, so one side of the cycles assignments is represented by
-the chord in the other.
-
-\mypic{simplechord}{There can not exists a cycle w-n-e-s without either
-a cycle n-s-w or n-e-s.}
-
-\Thomas{In mijn scriptie gebruik ik problem graphs en assignment graphs
-om de stap van undirected naar directed graphs te maken. Later kwam ik
-echter het begrip `orientation' tegen, met hetzelfde doel.}
-
-\section{Definitions}
-\input{./loags.tex}
-
-\DeclareRobustCommand{\VAN}[3]{#3}
-\bibliographystyle{splncs}
-\bibliography{biblio}
-
-\end{document}

### 2014-10-15 paper/paper.tex
Fix times in table (thanks Thomas!)

@@
       \hline
-      Kastens'                &   Y           &   25s \\
-      Kennedy-Warren  &   N           &   2s \\
-      LOAG-backtrack  &   N           &   35s \\
-      LOAG-SAT           &   N           &   10s \\
+      Kastens'                &   Y           &   17s \\
+      Kennedy-Warren  &   N           &   33s \\
+      LOAG-backtrack  &   N           &   36s \\
+      LOAG-SAT           &   N           &   9s \\
     \end{tabular}
@@
 In \tabref{tab:algo-comparison} we show the compilation times for the MainAG of the UHC for each of the four different algorithms.
-This time includes parsing of the attribute grammar description and code generation, which is what takes most time in the SAT approach.
+All times include parsing of the attribute grammar description and code generation, which is what takes most time in the SAT approach.
 The SAT-solver takes less than a second to find a solution.
-\Jeroen{De tijd voor Kennedy-Warren kan niet kloppen, Thomas kun jij die nog eens checken?
-Op mijn machine doen alle opties er veel langer over...}
 

### 2014-10-15 paper/paper.tex
Part on optimisations

@@
 \section{Optimisations} \label{sect:optimisations}
-\Jeroen{todo: laten zien dat we totaal aantal visits kunnen minimaliseren en dat we sommige attributen vroeg kunnen schedulen, en intuitie geven waarom dit nuttig is (meer hebben we ook nog niet)}
+We have shown that expressing the scheduling problem as a SAT problem and using an existing SAT-solver can improve the running time of the scheduling, but that is not the only advantage.
+In the SAT problem one can easily add extra constraints to further influence the resulting schedule.
+In this section we show two of such optimisations that are useful from an attribute grammar perspective.
+
+\subsection{Minimalising visits}
+The result of the static scheduling is a runtime evaluator that computes the values of the attributes for a given abstract syntax tree.
+The total order for each nonterminal defines in what order attributes should be computed, but in the implementation of the evaluator we make use of a slightly bigger unit of computation: a \emph{visit}.
+
+A visit is an evaluation step in the runtime evaluator that takes the values of a (possibly empty) set of inherited attributes and produces a (non-empty) set of synthesized attributes.
+In order to compute these values, visits to the child nodes may happen, and at the top of the tree the wrapping code invokes all visits of the top node one by one.
+
+Because invoking a visit at runtime may have a certain overhead, we would like the number of visits to be as small as possible.
+In other words, in the total order on the attributes we would like to minimise the number of places where a synthesized attribute is followed by an inherited attribute, because that is the location where a new visit needs to be performed.
+
+It is theoretically impossible to minimise the total number of visits performed for the full abstract syntax tree, because at compile-time we do not have a concrete abstract syntax tree at hand and only know about the grammar.
+We therefore try to minimise the maximum number of visits that any nonterminal takes.
+
+\Jeroen{todo: describe how}
+
+\subsection{Eager attributes}
+Another optimisation is the ability of defining \emph{eager attributes}.
+Eager attributes are attributes that should be computed as soon as possible, and must be annotated by the attribute grammar programmer as such.
+We would like our scheduling algorithm then to schedule them as early as possible in the total order.
+
+As an example, in a typical compiler there is an attribute containing the errors that occur in compilation.
+When running the compiler one is typically first interested in knowing if there are any errors; if so they must be printed to the screen and the compiler can stop its compilation.
+If there are no errors, then all other work that is not strictly necessary for the generation of errors can be done to complete the compilation.
+
+\Jeroen{todo: describe how}
 

### 2014-10-16 paper/paper.tex
Conclusion and discussion

@@
 \section{Conclusion and discussion} \label{sect:conclusion}
-\Jeroen{todo: het werkt, misschien is andere representatie beter, en het is onvoorspelbaarder dus misschien duren sommige ags ineens heel lang om te schedulen}
+We have explained the difficulties in attribute grammar scheduling and shown how to solve this problem using a SAT-solver.
+The given approach has been implemented and tested on a full-scale compiler built using attribute grammars.
+Results show that the algorithm works much faster than other approaches.
+
+In the translation to the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs.
+This technique makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions.
+We believe that this technique is applicable in other problems using SAT-solvers too.
+
+Furthermore, we have shown that expressing the problem as a SAT problem has the advantage that extra constraints can be added.
+We illustrated this with two possible properties of the resulting schedule that an attribute grammar programmer may want to influence.
+Even though this makes the scheduling problem potentially harder, as the algorithm is left fewer choices, the solution is found very fast in all practical cases we have tried.
+
+Another benefit of this approach that the attribute grammar scheduling can benefit from breakthroughs in the SAT community.
+The more efficient SAT solvers become, the better the attribute grammar scheduling becomes leading to larger and larger attribute grammars that are feasible to schedule.
+
+One problem with the current approach in constrast to most other scheduling algorithms is the unpredictability.
+SAT-solvers use certain heuristics to quickly find solutions in many applications, but it can theoretically happen that for a certain attribute grammar the SAT problem that is generated is not suitable for these heuristics.
+A seemingly innocent change in the attribute grammar definition could therefore theoretically lead to a large increase in compile time.
+However, we have not encountered this problem and we believe that this situation is very unlikely to happen because of the maturity of the SAT-solvers.
 

### 2014-10-16 paper/paper.tex
Remove some comments

@@
 Apart from the fact that this translation to SAT helps in efficiently (in practice) solving the scheduling problem, there also is another benefit: it is now possible to encode extra constraints on the resulting schedule.
-We show two \Jeroen{de 2e moet nog uitgewerkt worden} scheduling optimisations that are interesting from an attribute grammar perspective, for which the optimal schedule can be found efficiently by expressing the optimisation in the SAT problem.
+We show two scheduling optimisations that are interesting from an attribute grammar perspective, for which the optimal schedule can be found efficiently by expressing the optimisation in the SAT problem.
 
@@
 In order to encode this we also construct a \emph{nonterminal dependency graph}.
-This graph contains all attributes that are defined for the corresponding nonterminal and the edges in this graph must define a total order\Thomas{Actually it can be partial, which is also what we obtain using a variable for every inherited x synthesized pair.
-For example, an interface with only 1 visit (and multiple inherited or synthesized attributes) defines a partial order since we don't care about the order between inherited/synthesized attributes in the same visit.
-The visit-sequences that are calculated in the end do imply a total order however.}\Jeroen{Yes, but it can also just be total, which is much easier to explain, and I think this detail doesn't matter here} on the attributes.
+This graph contains all attributes that are defined for the corresponding nonterminal and the edges in this graph must define a total order on the attributes.
 Furthermore, the edges in the nonterminal dependency graph should agree with the (indirect) dependencies from the production dependency graphs such that no cycles exist.

### 2014-10-16 paper/paper.tex
Switch to author/year citations

@@
 \Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating to SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
-In our implementation we have used MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.
+In our implementation we use MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.
 
@@
 \section{Empirical results} \label{sect:results}
-As a test case for our work we use the Utrecht Haskell Compiler (UHC) \cite{Dijkstra:2009}.
+We use the Utrecht Haskell Compiler (UHC) \cite{Dijkstra:2009} as a test case for our work.
 The source code of the UHC consists of several attribute grammar definitions together with Haskell code.
@@
 However, the scheduling is not completely static so the generated evaluator also does part of the scheduling at runtime.
-\item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm \cite{Binsbergen:2015} for linear ordered attribute grammars, based on Kastens' algorithm.
+\item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm for linear ordered attribute grammars \cite{Binsbergen:2015}, based on Kastens' algorithm.
 This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper.
@@
 \DeclareRobustCommand{\VAN}[3]{#3}
-\bibliographystyle{splncs}
+\bibliographystyle{apalike}
 \bibliography{biblio}

### 2014-10-16 paper/paper.tex
Spellcheck

@@
 For a production two different types of attributes can be distinguished: \emph{input} attributes, which are attributes for which the value is provided, and \emph{output} attributes, which are values that need to be defined.
-The input attributes of a production, colored gray in the picture, are the inherited attributes of the production itself, because their values are provided by the parent node, and the synthesized attributes of all its children.
-The output attributes, colored white in the picture, are the synthesized attributes of the parent and the inherited attributes of the child.
+The input attributes of a production, coloured gray in the picture, are the inherited attributes of the production itself, because their values are provided by the parent node, and the synthesized attributes of all its children.
+The output attributes, coloured white in the picture, are the synthesized attributes of the parent and the inherited attributes of the child.
 
@@
 There the attribute grammar definitions are translated to lazy Haskell functions in a straightforward way by producing functions that take the inherited attributes as argument and return the synthesized attributes.
-Whenever an inherited attribute (indirectly) depends on the value of a synthesized attribute, this means the function has a cyclic definition, which is no problem for languages with lazy semantics, and can actually result in effient code \cite{Bird:1984}.
+Whenever an inherited attribute (indirectly) depends on the value of a synthesized attribute, this means the function has a cyclic definition, which is no problem for languages with lazy semantics, and can actually result in efficient code \cite{Bird:1984}.
 
@@
 However, these cycles could have been detected at compile time!
-Furthermore, the code needs lazy functions while with the static scheduling strict evaluators can be generated, leading to more effient code in practice.
+Furthermore, the code needs lazy functions while with the static scheduling strict evaluators can be generated, leading to more efficient code in practice.
 
@@
 The direct dependencies coming from the source code are constants, but for the rest of the edges the SAT solver can decide on the direction.
-However, the encoding has been chosen in such way that a valid assigment of the variables corresponds to a valid schedule.
+However, the encoding has been chosen in such way that a valid assignment of the variables corresponds to a valid schedule.
 
@@
 In this table we use three different sets: $\mathcal{D}$ is the set of direct dependencies, $\mathcal{C}$ is the set of edges that are added to make the graph chordal and $\mathcal{S}$ is the set of other edges.
-For each of the sets we take only the edges in the neigbourhood of \emph{v} for comparison.
+For each of the sets we take only the edges in the neighbourhood of \emph{v} for comparison.
 
@@
 
-\subsection{Minimalising visits}
+\subsection{Minimising visits}
 The result of the static scheduling is a runtime evaluator that computes the values of the attributes for a given abstract syntax tree.
@@
 
-One problem with the current approach in constrast to most other scheduling algorithms is the unpredictability.
+One problem with the current approach in contrast to most other scheduling algorithms is the unpredictability.
 SAT-solvers use certain heuristics to quickly find solutions in many applications, but it can theoretically happen that for a certain attribute grammar the SAT problem that is generated is not suitable for these heuristics.

### 2014-10-17 paper/paper.tex
Proofreading (improvements sections still missing)

@@
 Examples of compilers written using attribute grammars are JastAddJ \cite{Ekman:2007}, a Java compiler, and UHC \cite{Dijkstra:2009}, a Haskell compiler.
-The UHC is also the motivation for this paper and is used as a real world test case for testing the effectiveness of the given approach.
+The UHC is also the motivation for this paper and is used as a real world test case for evaluating the effectiveness of the given approach.
 
@@
 It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order.
-Such an evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps.
+Such an evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps.
 Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.
@@
 \item We show that chordal graphs can be used to encode cycle-freeness in SAT-problems involving graph encoding
-\item We show how certain attribute grammar optimisations can be encode in the SAT-problem
+\item We show how certain attribute grammar optimisations can be encoded in the SAT-problem
 \end{itemize}
@@
 
-In \figref{fig:labval_bin_flow_induced} we show the production dependency graph of the nonterminal \emph{Bin} with a complete order on the attributes.
+In \figref{fig:labval_bin_flow_induced} we show the production dependency graph of the production \emph{Bin} with a complete order on the attributes.
 In this case there is only a single nonterminal so both the parent and the child nodes have the same attribute order, made explicit by extra edges from the nonterminal dependency graph.
@@
 
+\subsection{Shared edges}
+One important implementation detail is that of shared edges.
+As explained, the nonterminal dependency graphs and the production dependency graphs share the edges that define the order of the attributes.
+Because each edge is represented by a variable in the SAT problem we can simply encode this by assigning the same variable to the shared edges.
+
+However, as we also make both graphs chordal, the edges added to make the graphs chordal can also be shared.
+This is exactly what our implementation does, such that the SAT-problem is kept as small as possible.
+The implementation is therefore slightly more complicated than explained in the previous sections.
+
 \section{Empirical results} \label{sect:results}
@@
 However, with extra manual annotations this algorithm can generate a schedule, which is the approach used previously.
-However, as writing these annotation is hard an error-prone, this is not an optimal solution.
+However, as writing these annotation is hard an error-prone, this is not an optimal workflow.
 \item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linear ordered attribute grammars.
@@
 \item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm for linear ordered attribute grammars \cite{Binsbergen:2015}, based on Kastens' algorithm.
-This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper.
+This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper and uses exponential time in worst-case.
 \end{itemize}
@@
 In this section we show two of such optimisations that are useful from an attribute grammar perspective.
+These optimisations have not been implemented in the release version of the UUAGC, but we have run preliminary experiments to verify that they work as expected.
 
@@
 \subsection{Eager attributes}
-Another optimisation is the ability of defining \emph{eager attributes}.
+Another optimisation is the ability to define \emph{eager attributes}.
 Eager attributes are attributes that should be computed as soon as possible, and must be annotated by the attribute grammar programmer as such.
@@
 This technique makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions.
-We believe that this technique is applicable in other problems using SAT-solvers too.
+We believe that this technique is applicable in other problems using SAT-solvers as well.
 
@@
 
+All in all, we believe that this approach fully solves the basic scheduling problem in an elegant way.
+There are ample possibilities for improving the resulting schedules based on attribute grammar knowledge like the two discussed in \sectref{sect:optimisations}, so we have also made room for future improvements in the scheduling and compilation of attribute grammars.
+
 \DeclareRobustCommand{\VAN}[3]{#3}

### 2014-10-17 paper/paper.tex
Optimisations

@@
 In the SAT problem one can easily add extra constraints to further influence the resulting schedule.
-In this section we show two of such optimisations that are useful from an attribute grammar perspective.
+In \sectref{sect:minimising} and \sectref{sect:eagerattributes} we show two of such optimisations that are useful from an attribute grammar perspective.
 These optimisations have not been implemented in the release version of the UUAGC, but we have run preliminary experiments to verify that they work as expected.
 
-\subsection{Minimising visits}
+\subsection{Interacting with the solver}
+Instead of directly expressing all constraints in the initial SAT problem, we use a different trick for implementing the two optimisations: interacting with the solver.
+After the initial scheduling problem has been solved, we can ask for the truth value of all variables to construct the schedule.
+MiniSat also keeps some state in memory that allows us to add extra constraints to the problem, and ask for a new solution.
+In this way we can start with an initial solution and interact with the solver until some optimum has been reached.
+
+\subsection{Minimising visits} \label{sect:minimising}
 The result of the static scheduling is a runtime evaluator that computes the values of the attributes for a given abstract syntax tree.
@@
 It is theoretically impossible to minimise the total number of visits performed for the full abstract syntax tree, because at compile-time we do not have a concrete abstract syntax tree at hand and only know about the grammar.
-We therefore try to minimise the maximum number of visits that any nonterminal takes.
+We therefore try to minimise the maximum number of visits for any nonterminal, which is the number of alternating pairs of inherited and synthesized attributes in the total order.
 
-\Jeroen{todo: describe how}
+We use the following algorithm:
+\begin{enumerate}
+\item Construct the initial SAT problem and solve
+\item Repeat while there is a solution:
+\begin{enumerate}
+\item Find the longest chain of alternating inherited/synthesized attributes (which is the maximum number of visits)
+\item If this maximum is lower then the previous best, store this result
+\item Add a constraint stating that at least one of the edges in this chain must be reversed
+\item Call solve again
+\end{enumerate}
+\end{enumerate}
 
-\subsection{Eager attributes}
+Remark that removing the longest chain of visits might lead to a worst schedule, as this could result in other nonterminals using even more visits.
+We therefore need to store the best result found so far separately.
+However, it is guaranteed that this procedure will find the optimal result, and experiments have shown it usually does so quickly.
+
+\subsection{Eager attributes} \label{sect:eagerattributes}
 Another optimisation is the ability to define \emph{eager attributes}.
@@
 
-\Jeroen{todo: describe how}
+We can use the following algorithm to implement this:
+\begin{enumerate}
+\item Construct the initial SAT problem and solve
+\item Create a set $\mathcal{E}$ with the eager attributes
+\item While $\mathcal{E}$ is not empty:
+\begin{enumerate}
+\item Take an attribute $\mathcal{A}$ from the set $\mathcal{E}$
+\item For all dependencies of $\mathcal{A}$, add constraint that at least one the edges must be reverted
+\item Solve, and if there is a solution:
+\begin{enumerate}
+\item Add $\mathcal{A}$ again to the set $\mathcal{E}$
+\end{enumerate}
+\item If there is no solution:
+\begin{enumerate}
+\item Revert to the previous solution (MiniSat allows this)
+\item Add all dependencies of $\mathcal{A}$ to $\mathcal{E}$
+\end{enumerate}
+\end{enumerate}
+\end{enumerate}
+
+One important remark here is that $\mathcal{E}$ does not necessarily contain attributes from the same nonterminal.
+If at the root of the abstract syntax tree we would like to have an attribute depend on as few other attributes as possible, it may be the case that deeper in the tree the attributes need to be reordered in order to achieve that goal.
+The dependencies of $\mathcal{A}$ therefore means the dependencies in a production nonterminal graph, which can include attributes of other nonterminals (because children of a production can be of different nonterminals).
+
+In contrast to the previous optimisation, this algorithm does not find a global optimum.
+Because optimistic choices are made this algorithm finds a local minimum, which may be much worse than the global optimum.
+We have not been able to verify what the influences of this are.
 

### 2014-10-17 paper/paper.tex
Some comments by Atze

@@
 \begin{abstract}
-Many computations over trees can be expressed using attribute
+Many computations over trees can be specified using attribute
 grammars. Compilers for attribute grammars need to find an evaluation
@@
 In this paper, we show how to encode linear ordered attribute grammar
-scheduling as a SAT-problem. Cycle-freeness of the dependency graphs
-is expressed by making the graphs chordal, which is a new approach for
-encoding such graph properties in SAT.
+scheduling as a SAT-problem.
+For such grammars it is necessary to ensure that the dependency graph is cycle free, which we approach in a novel way by transforming the dependency graph to a chordal graph allowing the cycle freeness to be efficiently expressed and computed using SAT solvers.
 
@@
 Examples of compilers written using attribute grammars are JastAddJ \cite{Ekman:2007}, a Java compiler, and UHC \cite{Dijkstra:2009}, a Haskell compiler.
-The UHC is also the motivation for this paper and is used as a real world test case for evaluating the effectiveness of the given approach.
+The use of attribute grammars in the UHC also motivates this paper and as such UHC is a real world test case for evaluating the effectiveness of the given approach.
 
@@
 It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order.
-Such an evaluation order or schedule should satisfy the dependencies from the attribute grammar definition and must be found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps.
+Such an evaluation order or schedule should satisfy the dependencies induced by the specification of attribute computations in the attribute grammar definition and must be found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps.
 Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.
@@
 \subsection{Summary}
-We solve the linear ordered attribute grammar scheduling problem by translating it to the Boolean satisfiability problem (SAT), one of the best-known NP-complete problems \cite{Cook:1971}.
+We solve the linear ordered attribute grammar scheduling problem by translating it into the Boolean satisfiability problem (SAT), one of the best-known NP-complete problems \cite{Cook:1971}.
 The SAT problem has been the subject of a lot of research and even though the worst case runtime of all known SAT-solving algorithms is exponential in the input size, many SAT-solvers work very well in practice.
-\Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating to SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
+\Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating into SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
 In our implementation we use MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.
@@
 
-Apart from the fact that this translation to SAT helps in efficiently (in practice) solving the scheduling problem, there also is another benefit: it is now possible to encode extra constraints on the resulting schedule.
+Apart from the fact that this translation into SAT helps in efficiently (in practice) solving the scheduling problem, there also is another benefit: it is now possible to encode extra constraints on the resulting schedule.
 We show two scheduling optimisations that are interesting from an attribute grammar perspective, for which the optimal schedule can be found efficiently by expressing the optimisation in the SAT problem.
@@
 We start in \sectref{sect:ag} and \sectref{sect:loag} by explaining linear ordered attribute grammars and the difficulties in scheduling in more detail.
-We then describe the translation to SAT in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}.
+We then describe the translation into SAT in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}.
 In \sectref{sect:optimisations} we describe the two attribute grammar optimisations and finally in \sectref{sect:conclusion} we discuss some problems and conclude.
@@
 For dynamic scheduling one could rely on existing machinery for lazy evaluation, for example in Haskell, which is the approach taken by \cite{saraiva99}.
-There the attribute grammar definitions are translated to lazy Haskell functions in a straightforward way by producing functions that take the inherited attributes as argument and return the synthesized attributes.
+There the attribute grammar definitions are translated into lazy Haskell functions in a straightforward way by producing functions that take the inherited attributes as argument and return the synthesized attributes.
 Whenever an inherited attribute (indirectly) depends on the value of a synthesized attribute, this means the function has a cyclic definition, which is no problem for languages with lazy semantics, and can actually result in efficient code \cite{Bird:1984}.
@@
 
-\section{Translation to SAT} \label{sect:translation}
+\section{Translation into SAT} \label{sect:translation}
 To represent the scheduling problem as a Boolean formula we introduce a variable for each edge, indicating the direction of the edge.
@@
 
-In the translation to the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs.
+In the translation into the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs.
 This technique makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions.

### 2014-10-17 paper/paper.tex
More comments by Atze

@@
 It is therefore up to the attribute grammar compiler (the compiler that translates the attribute grammar definition into a runtime evaluator) to find an evaluation order.
-Such an evaluation order or schedule should satisfy the dependencies induced by the specification of attribute computations in the attribute grammar definition and must be found statically, which means that for the given grammar an evaluation order should be found such that for every finite parse tree of the grammar the evaluation order should compute all values in a finite number of steps.
+Such an evaluation order or schedule should satisfy the dependencies induced by the specification of attribute computations in the attribute grammar definition and must be found statically, which means that for the given grammar an evaluation order should be found such that for every finite abstract syntax tree of the grammar the evaluation order should compute all values in a finite number of steps.
 Specifically, no cyclic data dependencies may occur at runtime for any given parse tree.
@@
 The SAT problem has been the subject of a lot of research and even though the worst case runtime of all known SAT-solving algorithms is exponential in the input size, many SAT-solvers work very well in practice.
-\Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating into SAT we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
+\Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating into the SAT problem we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
 In our implementation we use MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.
 
-The core of the scheduling problem is essentially a set of dependency graphs (directed acyclic graphs) for which a total order on the nodes should be found such that all direct dependencies, which are coming from the input grammar, are satisfied.
-However, there is a certain interplay between the different dependency graphs resulting in indirect dependencies, coming from the order that is found in one dependency graph, that need to be satisfied another dependency graph.
+The core of the scheduling problem consists of finding a total order on the nodes of a set of dependency graphs (directed acyclic graphs) such that all direct dependencies, which are coming from the input grammar, are satisfied.
+However, there is interplay between the different dependency graphs caused by the order that is found in one dependency graph resulting in indirect dependencies that need to be satisfied in another dependency graph.
 This interplay of dependencies, which is explained later in detail, is what makes the problem hard.
 
-To encode this problem in SAT we represent each edge as a Boolean variable, with its value indicating the direction of the edge.
+To encode this problem in SAT we represent each edge in the dependency graphs as a Boolean variable, with its value indicating the direction of the edge.
 For the direct dependencies the value is already set, but for the rest of the variables the SAT-solver may choose the direction of the edge.
-To ensure cycle-freeness this requires us to encode transitivity with SAT-constraints, which in the straight-forward solution leads to a number of extra constraints cubing in the number of variables.
-To avoid that problem we make our graphs \emph{chordal} \cite{Dirac:1961}, which means that for every cycle of size $> 3$, there must exist an edge between two non-adjacent nodes in the cycle.
+Ensuring cycle freeness requires us to encode transitivity with SAT-constraints, which in the straight-forward solution leads to a number of extra constraints cubic in the number of variables.
+To avoid that problem we make our graphs \emph{chordal} \cite{Dirac:1961}.
+In a chordal graph every cycle of size $> 3$ contains an edge between two non-adjacent nodes in the cycle.
 In other words, if there exists a cycle in the graph, there must also exist a cycle of at most three nodes.
-This allows us to encode cycle-freeness much more efficiently by only disallowing cycles of length three.
+This allows us to encode cycle freeness much more efficiently by only disallowing cycles of length three.
 
-Apart from the fact that this translation into SAT helps in efficiently (in practice) solving the scheduling problem, there also is another benefit: it is now possible to encode extra constraints on the resulting schedule.
-We show two scheduling optimisations that are interesting from an attribute grammar perspective, for which the optimal schedule can be found efficiently by expressing the optimisation in the SAT problem.
+Apart from the fact that this translation into the SAT problem helps in efficiently (in practice) solving the scheduling problem, there also is another benefit: it is now possible to encode extra constraints on the resulting schedule.
+We show two scheduling optimisations that are interesting from an attribute grammar point of view, for which the optimal schedule can be found efficiently by expressing the optimisation in the SAT problem.
 
@@
 \begin{itemize}
-\item We show how to encode the problem of scheduling linear ordered attribute grammars as a SAT-problem
-\item We show that chordal graphs can be used to encode cycle-freeness in SAT-problems involving graph encoding
-\item We show how certain attribute grammar optimisations can be encoded in the SAT-problem
+\item We show how to encode the problem of scheduling linear ordered attribute grammars as a SAT problem
+\item We show that chordal graphs can be used to encode cycle freeness in SAT problems
+\item We show how certain attribute grammar optimisations can be encoded as part of the formulation of the SAT problem
 \end{itemize}
-Furthermore, we have implement the described techniques in the UUAGC\footnote{\url{http://www.cs.uu.nl/wiki/HUT/AttributeGrammarSystem}} \cite{combinator-languages} and show that the technique works well in practice for the UHC.
+Furthermore, we have implemented the described techniques in the UUAGC\footnote{\url{http://www.cs.uu.nl/wiki/HUT/AttributeGrammarSystem}} \cite{combinator-languages} and show that the technique works well in practice for the UHC.
 
@@
 We start in \sectref{sect:ag} and \sectref{sect:loag} by explaining linear ordered attribute grammars and the difficulties in scheduling in more detail.
-We then describe the translation into SAT in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}.
+We then describe the translation into the SAT problem in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}.
 In \sectref{sect:optimisations} we describe the two attribute grammar optimisations and finally in \sectref{sect:conclusion} we discuss some problems and conclude.
@@
 \section{Attribute Grammars} \label{sect:ag}
-Formally an attribute grammar consists of three parts: a context-free grammar, a set of attribute definitions and a set of semantic functions.
+An attribute grammar consists of three parts: a context-free grammar, a set of attribute definitions and a set of semantic functions.
 Instead of a context-free grammar describing the concrete syntax of the language, we feel that it is more intuitive to visualise the grammar as describing the abstract syntax tree of the language.
@@
 
-In the rest of this paper we do stick to the usual attribute grammar terminology, so with a \emph{nonterminal} we mean a type, and a \emph{production} is a constructor of that type.
+In the rest of this paper we stick to the usual attribute grammar terminology, so with a \emph{nonterminal} we mean a type, and a \emph{production} is a constructor of that type.
 Furthermore, the name \emph{lhs} (short for left-hand side) refers to the constructor itself, while the children of a production have separate names.
@@
 The figure shows what is called the \emph{production dependency graphs}.
-For a production two different types of attributes can be distinguished: \emph{input} attributes, which are attributes for which the value is provided, and \emph{output} attributes, which are values that need to be defined.
+For a production two different types of attributes can be distinguished: \emph{input} attributes, which are attributes for which the value is provided, and \emph{output} attributes, which are values that need to be defined as part of the production relating parent and children.
 The input attributes of a production, coloured gray in the picture, are the inherited attributes of the production itself, because their values are provided by the parent node, and the synthesized attributes of all its children.
@@
 Although we talk about dependency graphs in the context of scheduling, we actually draw the edges in the other direction.
-The edges in the picture denote data flow, which from an attribute grammar perspective is much more intuitive.
+The edges in the picture represent data flow, which from an attribute grammar perspective is much more intuitive.
 For example, in the \emph{Bin} production the attribute a1 of the child \emph{r} is computed from a1 from the parent and a3 from the child \emph{l}.
-The edges are thus always directed from an input attribute to an output attribute, and the true dependency graph can be obtained by reversing all edges.
+The edges are thus always directed from an input attribute to an output attribute, and the actual dependency graph can be obtained by reversing all edges.
 
 \section{Linear Ordered Attribute Grammars} \label{sect:loag}
-Given an attribute grammar definition the attribute grammar compiler should produce a runtime evaluator that takes an abstract syntax tree as input and computes all attribute values.
+Given an attribute grammar definition the attribute grammar compiler generates a runtime evaluator that takes an abstract syntax tree as input and computes all attribute values.
 There are two approaches for doing the scheduling: at runtime (dynamic) or at compile time (static).
@@
 There are two problems with the dynamic scheduling.
-First of all whenever the attribute grammar contains a true cyclic attribute definition, the evaluator enters an infinite loop at runtime.
+First, whenever the attribute grammar contains a true cyclic attribute definition, the evaluator enters an infinite loop at runtime.
 However, these cycles could have been detected at compile time!
@@
 
-In \figref{fig:labval_bin_flow_induced} we show the production dependency graph of the production \emph{Bin} with a complete order on the attributes.
+\figref{fig:labval_bin_flow_induced} shows the production dependency graph of the production \emph{Bin} with a complete order on the attributes.
 In this case there is only a single nonterminal so both the parent and the child nodes have the same attribute order, made explicit by extra edges from the nonterminal dependency graph.
@@
 To represent the scheduling problem as a Boolean formula we introduce a variable for each edge, indicating the direction of the edge.
-The direct dependencies coming from the source code are constants, but for the rest of the edges the SAT solver can decide on the direction.
+The direct dependencies coming from the source code are constants, but for the rest of the edges the SAT-solver can decide on the direction.
 However, the encoding has been chosen in such way that a valid assignment of the variables corresponds to a valid schedule.
@@
 A chord is an edge between two nodes in the cycle that are not adjacent.
-The result of this is that each cycle of length $> 3$ is split up into two smaller cycles, so if a chordal graph contains a cycle it must also contain a cycle of size three.
+As a consequence each cycle of length $> 3$ can be split up into two smaller cycles, so if a chordal graph contains a cycle it must also contain a cycle of size three.
 Chordal graphs are therefore sometimes also referred to as \emph{triangulated graphs}.
@@
 However, as we also make both graphs chordal, the edges added to make the graphs chordal can also be shared.
-This is exactly what our implementation does, such that the SAT-problem is kept as small as possible.
+This is exactly what our implementation does, such that the SAT problem is kept as small as possible.
 The implementation is therefore slightly more complicated than explained in the previous sections.
@@
 However, with extra manual annotations this algorithm can generate a schedule, which is the approach used previously.
-However, as writing these annotation is hard an error-prone, this is not an optimal workflow.
+However, as writing these annotation is hard an error-prone, this is not an optimal work flow.
 \item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linear ordered attribute grammars.
@@
 \item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm for linear ordered attribute grammars \cite{Binsbergen:2015}, based on Kastens' algorithm.
-This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper and uses exponential time in worst-case.
+This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper and uses exponential time in worst case.
 \end{itemize}

### 2014-10-18 paper/paper.tex
Small fixes to the abstract and intro.

@@
 grammars. Compilers for attribute grammars need to find an evaluation
-order (or schedule) in order to generate efficient code. For the class
-of linear ordered attribute grammars such a schedule can be found
+order (or {\em schedule}) in order to generate efficient code. For the class
+of {\em linear ordered attribute grammars} such a schedule can be found
 statically, but this problem is known to be NP-hard.
@@
 Attribute Grammars \cite{knuth68} are a formalism for describing the semantics of context-free languages, thereby making them suitable for the construction of compilers.
-Examples of compilers written using attribute grammars are JastAddJ \cite{Ekman:2007}, a Java compiler, and UHC \cite{Dijkstra:2009}, a Haskell compiler.
-The use of attribute grammars in the UHC also motivates this paper and as such UHC is a real world test case for evaluating the effectiveness of the given approach.
+Examples of compilers written using attribute grammars are the Java compiler JastAddJ \cite{Ekman:2007}, and the Haskell compiler UHC \cite{Dijkstra:2009}.
+The use of attribute grammars in the UHC motivates this paper and as such UHC is a real world test case for evaluating the effectiveness of the given approach.
 
@@
 However, despite what the name suggests there exist attribute grammars for which a static evaluation order can be found that do not belong to the class of ordered attribute grammars.
-In our work on compiler construction we often encountered this situation, giving rise to the need of other scheduling algorithms \cite{bransen2012,Binsbergen:2015}.
+In our work on compiler construction we often encountered such grammars, giving rise to the need of other scheduling algorithms \cite{bransen2012,Binsbergen:2015}.
 We therefore look at the class of \emph{linear ordered attribute grammars} in this paper, which is the largest class of attribute grammars for which a static evaluation order can be found.
@@
 \subsection{Summary}
-We solve the linear ordered attribute grammar scheduling problem by translating it into the Boolean satisfiability problem (SAT), one of the best-known NP-complete problems \cite{Cook:1971}.
-The SAT problem has been the subject of a lot of research and even though the worst case runtime of all known SAT-solving algorithms is exponential in the input size, many SAT-solvers work very well in practice.
-\Jeroen{todo: referenties hier, misschien naar SAT-solving overzicht?} By translating into the SAT problem we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
+We solve the linear ordered attribute grammar scheduling problem by translating it into the Boolean satisfiability problem (SAT), the standard NP-complete problem \cite{Cook:1971}.
+Even though the worst case runtime of all known SAT-solving algorithms is exponential in the input size, many SAT-solvers work very well in practice \cite{sat-practice}.
+By translating into the SAT problem we can therefore use an efficient existing SAT-solver to solve our problem, and even benefit from future improvements in the SAT-community.
 In our implementation we use MiniSat\footnote{\url{http://minisat.se}} \cite{Een:2004}.
@@
 For the direct dependencies the value is already set, but for the rest of the variables the SAT-solver may choose the direction of the edge.
-Ensuring cycle freeness requires us to encode transitivity with SAT-constraints, which in the straight-forward solution leads to a number of extra constraints cubic in the number of variables.
+Ensuring cycle-freeness requires us to encode transitivity with SAT-constraints, which in the straight-forward solution leads to a number of extra constraints cubic in the number of variables.
 To avoid that problem we make our graphs \emph{chordal} \cite{Dirac:1961}.
@@
 \item We show how to encode the problem of scheduling linear ordered attribute grammars as a SAT problem
-\item We show that chordal graphs can be used to encode cycle freeness in SAT problems
+\item We show that chordal graphs can be used to encode cycle-freeness in SAT problems
 \item We show how certain attribute grammar optimisations can be encoded as part of the formulation of the SAT problem

### 2014-10-18 paper/paper.tex
Add a reference to Bryant's chordality paper.

@@
 This allows us to encode cycle freeness much more efficiently by only disallowing cycles of length three.
+Chordality has been used previously to encode equality logic in SAT using undirected graphs \cite{bryant-chordal}; to our knowledge this is the first application of chordality to express cycle-freeness of directed graphs.
 
@@
 
-\section{Conclusion and discussion} \label{sect:conclusion}
+\section{Discussion and Conclusion} \label{sect:conclusion}
 We have explained the difficulties in attribute grammar scheduling and shown how to solve this problem using a SAT-solver.

### 2014-10-18 paper/paper.tex
Add a ref to Bryant in the discussion as well.

@@
 In the translation into the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs.
-This technique makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions.
+The technique was inspired by existing work on chordility for expressing equality logic \cite{bryant-chordal}.
+Using chordal graphs makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions.
 We believe that this technique is applicable in other problems using SAT-solvers as well.

### 2014-10-18 paper/paper.tex
Rewrite optimisations to what they actually do.

@@
 
-We use the following algorithm:
+In our algorithm, we use efficient counting constraints, expressed in the SAT solver using sorting networks \cite{counting-sorting}.
+This enables us to count the number of true literals in a given set of literals, and express constraints about this number.
+A standard procedure for finding a solution for which the minimal number of literals in such a set is true can be implemented on top of a SAT-solver using a simple loop.
+
+We use the following algorithm for minimising the maximal number of visits:
 \begin{enumerate}
 \item Construct the initial SAT problem and solve
-\item Repeat while there is a solution:
+\item Construct the set of all production rules $P$
+\item Construct counting networks that count the number of visits $V(p)$ for all production rules $p$ in $P$
+\item Count the number of visits for each production rule in the current solution; let $M$ be the maximum value 
+\item Repeat while $M > 0$:
 \begin{enumerate}
-\item Find the longest chain of alternating inherited/synthesized attributes (which is the maximum number of visits)
-\item If this maximum is lower then the previous best, store this result
-\item Add a constraint stating that at least one of the edges in this chain must be reversed
-\item Call solve again
+\item Add constraints that express that for all productions $p$ in $P$: $V(p) \leq M$
+\item Construct a counting network that counts how many production rules $p$ have $V(p)=M$
+\item Compute a solution for which this number is minimised using the loop described above
+\item Remove all $p$ in $P$ for which now $V(p) = M$ from the set $P$
+\item Compute the new maximum value $M$ of $V(p)$ for all $p$ left in $P$
 \end{enumerate}
@@
 
-Remark that removing the longest chain of visits might lead to a worst schedule, as this could result in other nonterminals using even more visits.
-We therefore need to store the best result found so far separately.
-However, it is guaranteed that this procedure will find the optimal result, and experiments have shown it usually does so quickly.
+The above algorithm features a complicated combination of counting networks; one network for each production rule, and one network for each corresponding output of these networks.
+Still, the procedure finds optimal solutions very quickly in practice, in times that are negligable and not practically measurable compared to the time of generating the initial SAT-problem.
+The number of iterations for the minimisation loops has never been more than 5 in any of our problems.
+
+One could criticise the usefulness of this particular optimisation for attribute grammars.
+Indeed, details in how one should optimise the number of visits depend very much on the kind of trees we are going to run the compiled code on.
+Our point is that we can easily express variants of optimisations.
+For example, we can also minimise the sum of all visits using a similar (but simpler) procedure to the one above.
+Again, the running time of that procedure is very short.
 
@@
 
-We can use the following algorithm to implement this:
-\begin{enumerate}
-\item Construct the initial SAT problem and solve
-\item Create a set $\mathcal{E}$ with the eager attributes
-\item While $\mathcal{E}$ is not empty:
-\begin{enumerate}
-\item Take an attribute $\mathcal{A}$ from the set $\mathcal{E}$
-\item For all dependencies of $\mathcal{A}$, add constraint that at least one the edges must be reverted
-\item Solve, and if there is a solution:
-\begin{enumerate}
-\item Add $\mathcal{A}$ again to the set $\mathcal{E}$
-\end{enumerate}
-\item If there is no solution:
-\begin{enumerate}
-\item Revert to the previous solution (MiniSat allows this)
-\item Add all dependencies of $\mathcal{A}$ to $\mathcal{E}$
-\end{enumerate}
-\end{enumerate}
-\end{enumerate}
+In order to schedule a given attribute as early as possible, we are going to partition all attributes contained in the grammar into two sets $E$ (for early) and $L$ (for late).
+The idea is that $E$ contains all attributes that may be needed to be computed before the eager attribute (i.e.
+there exist production rules which require this), and $L$ contains all attributes that we can definitely compute after knowing the eager attribute (i.e.
+no production rule requires any attribute in $L$ for computing the eager attribute).
+We want to find a schedule for which the size of $E$ is minimal.
 
-One important remark here is that $\mathcal{E}$ does not necessarily contain attributes from the same nonterminal.
-If at the root of the abstract syntax tree we would like to have an attribute depend on as few other attributes as possible, it may be the case that deeper in the tree the attributes need to be reordered in order to achieve that goal.
-The dependencies of $\mathcal{A}$ therefore means the dependencies in a production nonterminal graph, which can include attributes of other nonterminals (because children of a production can be of different nonterminals).
+To compute this, we introduce a SAT variable $E(a)$ for every attribute $a$, that expresses whether or not $a$ is in $E$ or not.
+We set $E(a)$ to be true for the initial eager attribute $a$.
+We go over the graphs for the non-terminals and production rules, and generates constraints that express that whenever $a$ points to $b$ and we have $E(b)$, then we also need $E(a)$.
 
-In contrast to the previous optimisation, this algorithm does not find a global optimum.
-Because optimistic choices are made this algorithm finds a local minimum, which may be much worse than the global optimum.
-We have not been able to verify what the influences of this are.
+Finally, we use a counting network for all literals $E(a)$, and ask for a solution that minimises the number of literals in $E$.
+
+We have run this algorithm on ever output attribute of the top-level non-terminal of all our examples.
+For our largest grammar, the hardest output to compute took 1 second.
+So, while a harder optimisation than the previous one, it is very doable in practice.
+
+One can imagine other variants of this optimisation, where we have a set of eager attributes, or a combination of eager and late attributes.
+At the time of the writing of this paper, we have not experimented with such variants yet.
 

### 2014-10-18 paper/paper.tex
Add a statement about the number of visits found and how it related to previous known results.
Changed the paragraph about input and output attributes

@@
 
-The figure shows what is called the \emph{production dependency graphs}.
-For a production two different types of attributes can be distinguished: \emph{input} attributes, which are attributes for which the value is provided, and \emph{output} attributes, which are values that need to be defined as part of the production relating parent and children.
-The input attributes of a production, coloured gray in the picture, are the inherited attributes of the production itself, because their values are provided by the parent node, and the synthesized attributes of all its children.
-The output attributes, coloured white in the picture, are the synthesized attributes of the parent and the inherited attributes of the child.
+The figure shows \emph{production dependency graphs}, one for every production of the grammar.
+A set of semantic function definitions are given by the programmer for every production.
+The definitions specify how the synthesized attributes of a nonterminal are calculated in each of the different productions of that nonterminal.
+In semantic function definitions the inherited attributes of the nonterminal can be used as well as the synthesized attributes of its children.
+The attributes that are available to use, coloured gray in the picture, we call the \emph{input} attributes.
+Besides the semantic functions of the synthesized attributes of the parent node, the semantic functions of the inherited attributes of the children need to be defined.
+The attributes that require a definition, coloured white in the picture, we call the \emph{output} attributes.
+Note that with a semantic function definition for all output attributes we know how to compute all attributes as every input attribute is an output attribute of another node\footnote{Except for the inherited attributes of the root node.
+Its values need to be given as an argument to the semantic evaluator}.
 
@@
 The edges are thus always directed from an input attribute to an output attribute, and the actual dependency graph can be obtained by reversing all edges.
+The edges that follow directly from the source code are the \emph{direct dependencies}.
 
@@
 
+The algorithm is guaranteed to find the global optimum.
+For our largest example, the solution found had a total of 130 visits, which was 29 visits less in total than the previously known optimimum, found using backtracking heuristics.
+
 One could criticise the usefulness of this particular optimisation for attribute grammars.
@@
 We set $E(a)$ to be true for the initial eager attribute $a$.
-We go over the graphs for the non-terminals and production rules, and generates constraints that express that whenever $a$ points to $b$ and we have $E(b)$, then we also need $E(a)$.
+We go over the graphs for the nonterminals and production rules, and generates constraints that express that whenever $a$ points to $b$ and we have $E(b)$, then we also need $E(a)$.
 
@@
 
-We have run this algorithm on ever output attribute of the top-level non-terminal of all our examples.
+We have run this algorithm on ever output attribute of the top-level nonterminal of all our examples.
 For our largest grammar, the hardest output to compute took 1 second.
@@
 In the translation into the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs.
-The technique was inspired by existing work on chordility for expressing equality logic \cite{bryant-chordal}.
+The technique was inspired by existing work on chordality for expressing equality logic \cite{bryant-chordal}.
 Using chordal graphs makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions.

### 2014-10-18 paper/paper.tex
added a paragraph about the relation between nonterminal and production dependency graphs

@@
 Furthermore, the edges in the nonterminal dependency graph should agree with the (indirect) dependencies from the production dependency graphs such that no cycles exist.
+
+The nonterminal dependency graphs and production dependency graphs are consistent if 1) whenever there is a path from attribute $a$ to $b$ in a production dependency graph then there needs to be an edge from $a$ to $b$ in the nonterminal dependency graph in which $a$ and $b$ occur\footnote{There is no such nonterminal dependency graph if the two attributes belong to different nodes} and 2) if there is an edge from $a$ to $b$ in a nonterminal dependency graph then there needs to be an edge from $a$ to $b$ for all occurrences of $a$ and $b$ in the production dependency graphs.
 

### 2014-10-18 paper/paper.tex
Added a paragraph explaining which variables/edges we use in nonterminal graphs (WITH COMMENT THAT NEEDS TO BE REMOVED) and made which variables are added in the algorithm more explicit

@@
 
+To find the total order we add edges to the nonterminal dependency graph and rely on the SAT-solver to determine their direction.
+Instead of adding an edge for all possible pairs of attributes it is sufficient to add an edge for all pairs of inherited and synthesized attributes.
+The order described by an assignment to these variables is not (always) total as there might be pairs of attributes, which attributes are either both inherited or both synthesized, without a relative ordering.
+This is not a problem as their evaluation order can be chosen arbitrarily\Thomas{Dit beschouw ik zelf als een belangrijk resultaat.
+Het is de uitkomst van het lange proces waarin ik geprobeerd heb de minimale definitie van het probleem van LOAG scheduling te geven.
+Maar om dit beter te kunnen uitleggen heb ik 'visits' nodig.}.
+
 \section{Translation into SAT} \label{sect:translation}
@@
 \begin{enumerate}
-\item Construct a nonterminal dependency graph for each nonterminal
-\item Construct a production dependency graph for each production
+\item Construct a nonterminal dependency graph for each nonterminal and add an edge between all pairs of inherited and synthesized attributes
+\item Construct a production dependency graph for each production and add an edge for every direct dependency
 \item Make all graphs chordal
-\item Introduce a SAT variable for each edge in any of the graphs, where edges from a production dependency graph and the corresponding nonterminal dependency graph are shared, so are represented by the same variable
+\item Introduce a SAT variable for each edge in any of the graphs including the chords added in step 3.
+Variables of edges between attributes of the same nonterminal must be shared between nonterminal dependency graphs and production dependency graphs
 \item Set the value of all variables corresponding to direct dependencies
-\item Exclude all cycles of length three
+\item Exclude all cycles of length three by adding constraints
 \item Optionally add extra constraints for optimisations
@@
 In \tabref{tab:edgecomp} we show the results of different heuristics for the MainAG of the UHC.
-In this table we use three different sets: $\mathcal{D}$ is the set of direct dependencies, $\mathcal{C}$ is the set of edges that are added to make the graph chordal and $\mathcal{S}$ is the set of other edges.
+In this table we use three different sets: $\mathcal{D}$ is the set of direct dependencies (step 2, Section \ref{sect:translation}), $\mathcal{C}$ is the set of edges that are added to make the graph chordal (step 3, Section \ref{sect:translation}) and $\mathcal{S}$ is the set of edge between all inherited and synthesized pairs (step 1, Section \ref{sect:translation}).
 For each of the sets we take only the edges in the neighbourhood of \emph{v} for comparison.

### 2014-10-18 paper/paper.tex
kleine nuance
Last spellcheck, fix overfull hbox and remove last comment. This is the version to be submitted!

@@
 The order described by an assignment to these variables is not (always) total as there might be pairs of attributes, which attributes are either both inherited or both synthesized, without a relative ordering.
-This is not a problem as their evaluation order can be chosen arbitrarily\Thomas{Dit beschouw ik zelf als een belangrijk resultaat.
-Het is de uitkomst van het lange proces waarin ik geprobeerd heb de minimale definitie van het probleem van LOAG scheduling te geven.
-Maar om dit beter te kunnen uitleggen heb ik 'visits' nodig.}.
+This is not a problem as the evaluation order for such pairs can be chosen arbitrarily.
 
@@
 The above algorithm features a complicated combination of counting networks; one network for each production rule, and one network for each corresponding output of these networks.
-Still, the procedure finds optimal solutions very quickly in practice, in times that are negligable and not practically measurable compared to the time of generating the initial SAT-problem.
+Still, the procedure finds optimal solutions very quickly in practice, in times that are negligible and not practically measurable compared to the time of generating the initial SAT-problem.
 The number of iterations for the minimisation loops has never been more than 5 in any of our problems.
@@
 The algorithm is guaranteed to find the global optimum.
-For our largest example, the solution found had a total of 130 visits, which was 29 visits less in total than the previously known optimimum, found using backtracking heuristics.
+For our largest example, the solution found had a total of 130 visits, which was 29 visits less in total than the previously known optimum, found using backtracking heuristics.
 
@@
 In the translation into the SAT-problem we have used a novel technique for ruling out cycles in the SAT problem using chordal graphs.
-The technique was inspired by existing work on chordality for expressing equality logic \cite{bryant-chordal}.
+Existing work on chordality for expressing equality logic \cite{bryant-chordal} was the inspiration for this technique.
 Using chordal graphs makes the problems much smaller than directly ruling out cycles, while encoding the same restrictions.

### 2015-01-06 paper/paper.tex
paper: linear ordered AGs -> linearLY ordered AGs

@@
 % Paper info
-\title{Linear Ordered Attribute Grammar scheduling using SAT-solving}
+\title{Linearly Ordered Attribute Grammar scheduling using SAT-solving}
 \author{Jeroen Bransen\inst{1}  \and L. Thomas van Binsbergen\inst{2,1} \and Koen Claessen\inst{3} \and Atze Dijkstra\inst{1}}
@@
 order (or {\em schedule}) in order to generate efficient code. For the class
-of {\em linear ordered attribute grammars} such a schedule can be found
+of {\em linearly ordered attribute grammars} such a schedule can be found
 statically, but this problem is known to be NP-hard.
 
-In this paper, we show how to encode linear ordered attribute grammar
+In this paper, we show how to encode linearly ordered attribute grammar
 scheduling as a SAT-problem.
@@
 In our work on compiler construction we often encountered such grammars, giving rise to the need of other scheduling algorithms \cite{bransen2012,Binsbergen:2015}.
-We therefore look at the class of \emph{linear ordered attribute grammars} in this paper, which is the largest class of attribute grammars for which a static evaluation order can be found.
+We therefore look at the class of \emph{linearly ordered attribute grammars} in this paper, which is the largest class of attribute grammars for which a static evaluation order can be found.
 The problem of statically finding such evaluation order is known to be NP-complete \cite{engelfriet82}.
@@
 \subsection{Summary}
-We solve the linear ordered attribute grammar scheduling problem by translating it into the Boolean satisfiability problem (SAT), the standard NP-complete problem \cite{Cook:1971}.
+We solve the linearly ordered attribute grammar scheduling problem by translating it into the Boolean satisfiability problem (SAT), the standard NP-complete problem \cite{Cook:1971}.
 Even though the worst case runtime of all known SAT-solving algorithms is exponential in the input size, many SAT-solvers work very well in practice \cite{sat-practice}.
@@
 \begin{itemize}
-\item We show how to encode the problem of scheduling linear ordered attribute grammars as a SAT problem
+\item We show how to encode the problem of scheduling linearly ordered attribute grammars as a SAT problem
 \item We show that chordal graphs can be used to encode cycle-freeness in SAT problems
@@
 The outline of the paper is as follows.
-We start in \sectref{sect:ag} and \sectref{sect:loag} by explaining linear ordered attribute grammars and the difficulties in scheduling in more detail.
+We start in \sectref{sect:ag} and \sectref{sect:loag} by explaining linearly ordered attribute grammars and the difficulties in scheduling in more detail.
 We then describe the translation into the SAT problem in \sectref{sect:translation} and show the effectiveness of our approach in \sectref{sect:results}.
@@
 
-\section{Linear Ordered Attribute Grammars} \label{sect:loag}
+\section{Linearly Ordered Attribute Grammars} \label{sect:loag}
 Given an attribute grammar definition the attribute grammar compiler generates a runtime evaluator that takes an abstract syntax tree as input and computes all attribute values.
@@
 In \cite{kastens80} a polynomial time algorithm is given for static scheduling of the class of ordered attribute grammars, but unfortunately we often encounter attribute grammars outside of that class.
-All our attribute grammars do however fall in the class of linear ordered attribute grammars, which is the largest class of attribute grammars for which a static schedule can be found.
+All our attribute grammars do however fall in the class of linearly ordered attribute grammars, which is the largest class of attribute grammars for which a static schedule can be found.
 Although the scheduling is NP-complete, we have implemented a backtracking algorithm in earlier work \cite{Binsbergen:2015} that is feasible in practice.
@@
 
-Conceptually, static scheduling of linear ordered attribute grammars is not very complex.
+Conceptually, static scheduling of linearly ordered attribute grammars is not very complex.
 For each nonterminal a total order should be found on its attributes, such that all production dependency graphs all cycle free.
@@
 However, as writing these annotation is hard an error-prone, this is not an optimal work flow.
-\item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linear ordered attribute grammars.
+\item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linearly ordered attribute grammars.
 However, the scheduling is not completely static so the generated evaluator also does part of the scheduling at runtime.
-\item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm for linear ordered attribute grammars \cite{Binsbergen:2015}, based on Kastens' algorithm.
+\item \emph{LOAG-backtrack}: we have also implemented a backtracking algorithm for linearly ordered attribute grammars \cite{Binsbergen:2015}, based on Kastens' algorithm.
 This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper and uses exponential time in worst case.

### 2015-01-06 paper/paper.tex
Fix small typo
Define neighbourhood
Changed my email + some changes + change of LOAG-backtrack compiletime

@@
 \institute{Utrecht University, Utrecht, The Netherlands, \email{\{J.Bransen,atze\}@uu.nl}
-\and Royal Holloway, University of London, Egham, UK, \email{Thomas.VanBinsbergen.2014@live.rhul.ac.uk}
+\and Royal Holloway, University of London, Egham, UK, \email{ltvanbinsbergen@acm.org}
 \and Chalmers University of Technology, Gothenburg, Sweden, \email{koen@chalmers.se}}
@@
 \begin{definition}
-An undirected graph is \emph{chordal} if and only if it has a 
+An undirected graph $G$ is \emph{chordal} if and only if it has a 
 perfect elimination order. A perfect elimination order is an
@@
 the neighbourhood of $v$ forms a connected component in $G$.
+The neighbourhood of $v$ are all vertices $w$ such that there exists an edge between $v$ and $w$.
 The graph $G[v_1,\ldots,v_i]$ is the induced subgraph of $G$ containing
@@
 \subsection{Finding the schedule}
-When the constructed Boolean formula is given to the SAT-solver, the result is either that the formula is not satisfiable, meaning that no schedule has been found, of satisfiable, meaning that there is a schedule.
+When the constructed Boolean formula is given to the SAT-solver, the result is either that the formula is not satisfiable, meaning that no schedule can been found, or satisfiable, meaning that there is a schedule.
 It is not hard to see that the formula is satisfiable if and only if there exists a valid schedule for the given attribute grammar definition, and in this paper we give no formal proof of this claim.
@@
 However, with extra manual annotations this algorithm can generate a schedule, which is the approach used previously.
-However, as writing these annotation is hard an error-prone, this is not an optimal work flow.
+However, as writing these annotation is hard and error-prone, this is not an optimal work flow.
 \item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linearly ordered attribute grammars.
@@
 This algorithm solves exactly the same scheduling problem as the SAT approach described in this paper and uses exponential time in worst case.
+In case of the \emph{MainAG} no backtracking is required to find a schedule.
 \end{itemize}
@@
       Kennedy-Warren  &   N           &   33s \\
-      LOAG-backtrack  &   N           &   36s \\
+      LOAG-backtrack  &   N           &   13s \\
       LOAG-SAT           &   N           &   9s \\
@@
 In \tabref{tab:algo-comparison} we show the compilation times for the MainAG of the UHC for each of the four different algorithms.
-All times include parsing of the attribute grammar description and code generation, which is what takes most time in the SAT approach.
+All times include parsing of the attribute grammar description, code generation and adding chords, which is what takes most time in the SAT approach.
 The SAT-solver takes less than a second to find a solution.

### 2015-01-15 paper/paper.tex
Kleine typefout
Remove some 'much' and 'very' instances from paper

@@
 There are two main advantages to using a SAT-solver for scheduling:
-(1) the scheduling algorithm runs much faster than existing scheduling
+(1) the scheduling algorithm runs faster than existing scheduling
 algorithms on real-world examples, and (2) by adding extra constraints
@@
 In other words, if there exists a cycle in the graph, there must also exist a cycle of at most three nodes.
-This allows us to encode cycle freeness much more efficiently by only disallowing cycles of length three.
+This allows us to encode cycle freeness more efficiently by only disallowing cycles of length three.
 Chordality has been used previously to encode equality logic in SAT using undirected graphs \cite{bryant-chordal}; to our knowledge this is the first application of chordality to express cycle-freeness of directed graphs.
@@
 
-We have run this algorithm on ever output attribute of the top-level nonterminal of all our examples.
+We have run this algorithm on every output attribute of the top-level nonterminal of all our examples.
 For our largest grammar, the hardest output to compute took 1 second.
@@
 The given approach has been implemented and tested on a full-scale compiler built using attribute grammars.
-Results show that the algorithm works much faster than other approaches.
+Results show that the algorithm works faster than other approaches.
 
@@
 A seemingly innocent change in the attribute grammar definition could therefore theoretically lead to a large increase in compile time.
-However, we have not encountered this problem and we believe that this situation is very unlikely to happen because of the maturity of the SAT-solvers.
+However, we have not encountered this problem and we believe that this situation is unlikely to happen because of the maturity of the SAT-solvers.
 

### 2015-01-16 paper/paper.tex
Tabel

@@
   \begin{center}
-    \begin{tabular}{l || c || r}
-      Algorithm   &   Manual annotations   &  Compile time MainAG \\
+    \begin{tabular}{l || c || r || r || r || r}
+      Algorithm   &   Kastens' & Kennedy-Warren & LOAG-backtrack & LOAG-SAT\\
       \hline
-      Kastens'                &   Y           &   17s \\
-      Kennedy-Warren  &   N           &   33s \\
-      LOAG-backtrack  &   N           &   13s \\
-      LOAG-SAT           &   N           &   9s \\
+      UHC's MainAG          &   -       & 33s   & 13s   & 9s    \\
+      Asil Test             &   -       & 1.8s  & 4.4s  & 3.4s  \\
+      Asil ByteCode         &   -       & 0.6s  & 29.4s & 2.8s  \\
+      Asil PrettyTree       &   -       & 390ms & 536ms & 585ms \\
+      Asil InsertLabels     &   -       & 314ms & 440ms & 452ms \\
+      UUAGC's CodeGeneration&   -       & 348ms & 580ms & 382ms \\
+      5 pigeonholes         &   -       & 107ms & 1970ms& 191ms \\ 
+      Helium's TS\_Analyse  &   190ms   & 226ms & 235ms & 278ms \\
     \end{tabular}

### 2015-01-16 paper/paper.tex
Final version of the paper

--- 2015-01-15 paper/paper.tex
@@
 
-There are two problems with the dynamic scheduling.
-First, whenever the attribute grammar contains a true cyclic attribute definition, the evaluator enters an infinite loop at runtime.
+There are two problems with dynamic scheduling.
+First, whenever the attribute grammar contains a true cyclic definition, the evaluator enters an infinite loop at runtime.
 However, these cycles could have been detected at compile time!
@@
 \section{Empirical results} \label{sect:results}
-We use the Utrecht Haskell Compiler (UHC) \cite{Dijkstra:2009} as a test case for our work.
+We use the Utrecht Haskell Compiler (UHC) \cite{Dijkstra:2009} as the main test case for our work.
 The source code of the UHC consists of several attribute grammar definitions together with Haskell code.
@@
 
-The UHC is compiled using the Utrecht University Attribute Grammar Compiler (UUAGC) \cite{combinator-languages}, in which we have implemented our approach.
+We have compiled several attribute grammars using the Utrecht University Attribute Grammar Compiler (UUAGC) \cite{combinator-languages}, in which we have implemented our approach.
 Apart from our SAT approach there are three other scheduling algorithms that have been implemented to which we can compare our approach:
@@
 \begin{itemize}
-\item \emph{Kastens}: the algorithm from \cite{kastens80} which only works for ordered attribute grammars, of which MainAG is not a member.
-However, with extra manual annotations this algorithm can generate a schedule, which is the approach used previously.
-However, as writing these annotation is hard and error-prone, this is not an optimal work flow.
+\item \emph{Kastens}: the algorithm from \cite{kastens80} which only works for ordered attribute grammars, of which most examples are not a member.
 \item \emph{Kennedy-Warren}: The algorithm from \cite{kennedywarren76} that we have implemented in the UUAGC before \cite{bransen2012} is an algorithm that schedules all absolutely noncircular attribute grammars, an even larger class than the linearly ordered attribute grammars.
@@
   \begin{center}
-    \begin{tabular}{l || c || r}
-      Algorithm   &   Manual annotations   &  Compile time MainAG \\
+    \begin{tabular}{l || c || r || r || r || r}
+      Algorithm   &   Kastens' & Kennedy-Warren & LOAG-backtrack & LOAG-SAT\\
       \hline
-      Kastens'                &   Y           &   17s \\
-      Kennedy-Warren  &   N           &   33s \\
-      LOAG-backtrack  &   N           &   13s \\
-      LOAG-SAT           &   N           &   9s \\
+      UHC MainAG          &   -       & 33s   & 13s   & 9s    \\
+      Asil Test             &   -       & 1.8s  & 4.4s  & 3.4s  \\
+      Asil ByteCode         &   -       & 0.6s  & 29.4s & 2.8s  \\
+      Asil PrettyTree       &   -       & 390ms & 536ms & 585ms \\
+      Asil InsertLabels     &   -       & 314ms & 440ms & 452ms \\
+      UUAGC CodeGeneration&   -       & 348ms & 580ms & 382ms \\
+      Pigeonhole principle         &   -       & 107ms & 1970ms& 191ms \\ 
+      Helium TS\_Analyse  &   190ms   & 226ms & 235ms & 278ms \\
     \end{tabular}
@@
 
-In \tabref{tab:algo-comparison} we show the compilation times for the MainAG of the UHC for each of the four different algorithms.
+In \tabref{tab:algo-comparison} we show the compilation times for the examples for the four different algorithms.
 All times include parsing of the attribute grammar description, code generation and adding chords, which is what takes most time in the SAT approach.
-The SAT-solver takes less than a second to find a solution.
+The SAT-solver takes less than a second to find a solution in all cases.
+
+The other test cases we have used for testing are the UUAGC itself, \emph{Asil} \cite{asil} which is a byte code instrumenter, the \emph{Helium} compiler \cite{helium} which is a Haskell compiler specifically intended for students learning Haskell, and an encoding of the Pigeonhole principle.
+We removed one clause in such way there exists exactly one valid schedule, resulting in an artificial attribute grammar that is hard to schedule.
 

### 2015-01-17 code/SAT2AG.hs
Update code a bit to take arguments to create a solvable instance

@@ -91,5 +91,6 @@ genAll cnf = genVars cnf >> genSat cnf >> genTop cnf
 
 main :: IO ()
 main = do
-  [n] <- getArgs
-  genAll $ pidgeon $ read n
+  args <- getArgs
+  let (n:ns) = map read args
+  genAll $ map snd $ filter (\(i,c) -> i `notElem` ns) $ zip [1..] $ pidgeon n
